---
title: "Term"
author: "Victor Yuan"
date: "27/05/2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

# 1.0 Setup

## Load libraries and data

```{r message = F, warning = F}
# libraries and data
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggrepel)
library(viridis)
library(scales)
library(RColorBrewer)
library(impute)
library(pheatmap)
library(irlba)
library(yahew)
library(GGally)
library(cowplot)
library(purrr)
library(tibble)
library(umap)
library(dendextend)
library(stringr)

# pdata
pDat <- readRDS('../../data/main/interim/13_pDat.rds')

# raw methylation data
betas <- readRDS('../../data/main/interim/14_betas_noob_filt.rds')
probe_anno <- readRDS('../../data/main/interim/11_probe_anno.rds')

# snp data
snp_betas <- readRDS('../../data/main/interim/11_snp_betas.rds')

# annotation
zhou_anno <- readRDS('Z:/Victor/Data/DNAm annotations/zhou2017_EPIC.hg19.manifest.rds') 

# color key
color_code <- readRDS('../../data/main/interim/11_color_code.rds')
color_code_tissue <- setNames(color_code[[1]]$Colors_Tissue, color_code[[1]]$Tissue)
```

## Filter to term

```{r}
dim(betas) #737050  192
table(pDat$Trimester) # 126 term samples.

all(colnames(betas) == pDat$Sentrix) # verify betas and pDat sample order match
colnames(betas) <- pDat$Sample_Name # input sample names as columna nmaes

pDat %>% filter(Trimester == 'Third') %>% group_by(Tissue) %>% count()
```

Some samples will we ran for secondary analysis, e.g. mixtures, maternal villi, and dead cells. 
These I will filter out and save for later

I keep syncytiotrophoblasts for now, since we have a good number of them (n=5), and should be 
relatively clean. Sample number is probably not sufficient for linear modelling though.

```{r}
pDat_1 <- pDat %>% 
  filter(Trimester == 'Third', 
         !Tissue %in% c('Dead Cells and Lymphocytes', 'Mixture', 'Villi maternal'),
         Case_ID != 'PM324')

betas_1 <- betas[,pDat_1$Sample_Name]

nrow(pDat_1) # 105 samples

table(pDat_1$Tissue)
```

# 2.0 Sample characteristics

Let's look at various sample characteristics

```{r}
glimpse(pDat_1)

cov_tests <- pDat_1 %>%
  select(Row_numeric, Row_factor, Row_numeric,
         Case_ID, Sex, Week, Tissue, Batch_BSC, failed_probes,
         Average_intensity,
         DNA_loaded,
         cor_to_reference, cor_to_unrelated, Prob_SNP_outlier) %>%
  as.data.frame() %>% pairtest

# make categories
cov_tests <- cov_tests %>% 
  mutate(pval_cat = if_else(p.value < 0.001, '< 0.001',
                            if_else(p.value < 0.01, '< 0.01',
                                    if_else(p.value < 0.05, '< 0.05', '> 0.05'))))

# plot heatmap of associations
ggplot(cov_tests, aes(x=Row, y = Column, fill = pval_cat)) +
  geom_tile(col = 'grey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = c('> 0.05' = 'White', '< 0.05' = '#fee8c8', 
                               '< 0.01' = '#fdbb84', '< 0.001' = '#e34a33')) +
  labs(x = '', y = '', fill = 'p') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust= 1),
        panel.grid.major.y = element_blank()) +
  coord_equal()

lm(pDat_1$Row_numeric ~ pDat_1$Average_intensity) %>% summary
lm(pDat_1$DNA_loaded ~ pDat_1$failed_probes) %>% summary
```

Not surprising of the following:

`Sex` - `Case ID`           * Because each case has to be only one sex
`Batch_BSC` - `Week `       * Because each BSC batch was done in a particular week
`Case ID` and `correlation to unrelated` makes sense given that all samples in the same case are
highly correlated at SNP probes, so their correlation to others should be highly similar.

Below I examine the distributions of the other significantly associated covariates:

```{r}
g1 <- ggplot(pDat_1, aes(y = cor_to_unrelated, x = Sex)) +
  geom_violin() + geom_jitter() + theme_bw() +labs(x = '')

g2 <- ggplot(pDat_1, aes(y = Prob_SNP_outlier, x = Tissue)) +
  geom_boxplot() + geom_jitter() + theme_bw() + labs(x = '') +
  theme(axis.text.x = element_text(angle =45, hjust = 1, vjust = 1))


g3 <- ggplot(pDat_1, aes(x = Row_numeric, y = Average_intensity)) +
  geom_point(alpha = 0.5, position = position_dodge2(0.1)) + theme_bw() + 
  geom_smooth(method = 'loess') + 
  labs(title = 'Rsquared = 0.15')

g4 <- ggplot(pDat_1, aes(x = DNA_loaded, y = failed_probes)) +
  geom_smooth(method = 'lm') + labs(title = 'Rsquared = 0.05', x = 'DNA loaded (ng)') +
  geom_jitter(alpha = 0.5, width = 1) + theme_bw()

plot_grid(g1, g2, g3, g4, ncol = 2)

ggplot(pDat_1, aes(x = Row_numeric, y = failed_probes)) +
  geom_point()
```

Importantly we see that Tissue is not confounded by any of our other variables.

# 3.0 Global methylation

## Density distributions

Here we plot the density distributions of the betas across each tissue

```{r}
# pull out sample names for each tissue
list <- list()
for (i in unique(pDat_1$Tissue)) {
  list[[i]] <- pDat_1 %>% filter(Tissue == i) %>% pull(Sample_Name)
}

# for each tissue estimate the density
densities <- tibble(Tissue = names(list),
       betas = lapply(list, function(x) as.vector(betas_1[,x]))) %>%
  mutate(densities = map(betas, ~ density(.))) %>%
  
   # pull out x and y coordinates
  mutate(x = map(densities, 'x'),
         y = map(densities, 'y')) %>%

  # remove input data, and unnest
  select(-betas, -densities) %>%
  unnest()

# plot
ggplot(densities, 
       aes(x = x, y = y, color = Tissue)) +
  geom_line(size = 2, alpha = 1) + theme_bw() +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1)) +facet_wrap(~Tissue) +
  scale_color_manual(values= color_code_tissue[unique(pDat_1$Tissue)])
```

Here I plot the median beta value for each cpg for each tissue

```{r}
med_densities <- tibble(Tissue = names(list),
                        
                        # compute medians across every cpg for each tissue
                        betas = lapply(list, function(x) apply(betas_1[,x], 1, median))) %>% 
  
  # compute density line over medians
  mutate(densities = map(betas, ~ density(.))) %>%
  
  # pull out x and y coordinates for density
  mutate(x = map(densities, 'x'),
         y = map(densities, 'y')) %>%
  
  # unlist
  select(-betas, -densities) %>%
  unnest()

ggplot(med_densities, aes(x = x, y = y, color = Tissue)) +
  geom_line(size = 2, alpha = 0.6) + theme_bw() +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1))  +
  scale_color_manual(values= color_code_tissue[unique(pDat_1$Tissue)])
```

For each sample:

```{r}
sample_densities <- pDat_1 %>%
  select(Sample_Name, Tissue) %>%
  mutate(densities = apply(betas_1, 2, density)) %>%
  mutate(x = map(densities, 'x'),
         y = map(densities, 'y')) %>%
  select(-densities) %>%
  unnest()

ggplot(sample_densities, aes(x = x, y = y, color = Tissue, group = Sample_Name)) +
  geom_line(size = 1, alpha = 0.5) + theme_bw() +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1))  +
  facet_wrap(~Tissue) +
  scale_color_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  geom_text_repel(data = sample_densities %>% 
                    filter(Sample_Name == 'PM364_hofb_cs', x > 0.55, x < 0.6) %>%
                    dplyr::slice(1),
                  aes(x = x, y = y, label = Sample_Name), inherit.aes = F,
                  nudge_y = 1, nudge_x = -0.1) +
  labs(x = '% methylation', y = 'density')
```

## PCA

```{r}
# compute pca
set.seed(1)
pca_1 <- prcomp_irlba(t(betas_1), n = 20, center = T, scale = F)

# add pc scores to pdata
pca_scores <- pca_1$x[,1:20] %>% as.data.frame()
colnames(pca_scores) <- paste0(colnames(pca_scores), '_processed')
pDat_1 <- pDat_1 %>% bind_cols(pca_scores) %>% select(-contains('_raw'))

# create proportion variance explained data frame
pc_info <- summary(pca_1)$importance %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = 'variable') %>%
  gather(key = 'PC', value = 'value', -variable) %>%
  as_tibble() %>%
  mutate(PC = factor(as.character(PC), levels = paste0('PC', 1:20)),
         Label = ifelse(variable == 'Proportion of Variance',
                        paste0(PC, ' (', prettyNum(value*100, digits = 2), '%)'),
                        as.character(PC))) %>%
  arrange(variable, PC)

  
# correlate PCs with phenodata
pc_cor <- lmmatrix(dep = pca_1$x[,1:20],
                   ind = pDat_1 %>%
                   dplyr::select(Case_ID, Tissue, Sex, #bio
                                 Week, Chip_number, Row_numeric, Row_factor, Batch_BSC, # batch
                                 DNA_loaded, 
                                 failed_probes,
                                 cor_to_reference, Prob_SNP_outlier),
                   metric = 'Pvalue')
# plot data
pc_cor <- pc_cor %>% as.data.frame() %>% 
  
  # add dep variables
  mutate(dep = rownames(pc_cor)) %>%
  
  # reshape
  gather(PC, pval, -dep) %>%
  
  # pvalue categories
  mutate(pval_cat = factor(case_when(
    pval > 0.05  ~ '> 0.05',
    pval < 0.05 & pval > 0.01 ~ '< 0.05',
    pval < 0.01 & pval > 0.001 ~ '< 0.01',
    pval < 0.001 ~ '< 0.001'
  ), levels = c('> 0.05', '< 0.05','< 0.01', '< 0.001')),
  
  # make PC is encoded with proper levels!!!
  PC = factor(PC, levels = paste0('PC', 1:20))) %>% as_tibble()

# create color palette
colpal <- c('white', '#fee8c8', '#fdbb84', '#e34a33')
names(colpal) <- levels(pc_cor$pval_cat)

p1 <- ggplot(pc_cor, aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile(col = 'lightgrey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0), labels = 1:20) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = colpal)  + 
  labs(y = '', fill = 'P value')

p2 <- ggplot(pc_info %>% filter(variable == 'Proportion of Variance') %>%
               mutate(value = value*100), 
             aes(x = PC, y = value)) +
  geom_bar(stat = 'identity') +
  theme_bw() + 
  scale_x_discrete(expand = c(0, 0), labels = 1:20) + 
  scale_y_continuous(expand = c(0, 0)) +
  labs(y = '% variance explained')

egg::ggarrange(p1, p2, heights = c(3,1))
```

Plot some scatterplots of significantly associated variables

```{r}
scatter <- function(x, y, fill, point_size = 1){
  xlab <- pc_info %>% filter(variable == 'Proportion of Variance', PC == x) %>% pull(Label)
  ylab <- pc_info %>% filter(variable == 'Proportion of Variance', PC == y) %>% pull(Label)
  
  x <- paste0(x, '_processed')
  y <- paste0(y, '_processed')
  
  out <- ggplot(pDat_1, aes_string(x = x, y = y, fill = fill)) +
    geom_point(shape = 21, size = point_size) + theme_bw() + labs(x = xlab, y = ylab) 
  
  if (is.numeric(as.data.frame(pDat_1)[,fill])){
    out <- out +
      scale_fill_viridis()
  } else {
    out <- out + 
      scale_fill_brewer(palette = 'Set1')
  }
    
  out
}
scatter(x = 'PC1', y = 'PC2', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  geom_text_repel(data = pDat_1 %>% filter((Tissue == 'Hofbauer' & PC1_processed < 0)|
                                             Case_ID == 'PM364'),
                  aes(x = PC1_processed, y = PC2_processed, label =Sample_Name),
                  nudge_y = 15, nudge_x = -5)
scatter(x = 'PC1', y = 'PC2', fill = 'cor_to_reference', point_size = 2)
scatter(x = 'PC1', y = 'PC2', fill = 'Prob_SNP_outlier', point_size = 2)
scatter(x = 'PC1', y = 'PC2', fill = 'DNA_loaded', point_size = 2)
# calculate association after adjusting for tissue effects
scatter(x = 'PC3', y = 'PC4', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)])
scatter(x = 'PC5', y = 'PC6', fill = 'failed_probes', point_size = 2)
scatter(x = 'PC5', y = 'PC6', fill = 'DNA_loaded', point_size = 2)
scatter(x = 'PC5', y = 'PC6', fill = 'Week', point_size = 2)
scatter(x = 'PC5', y = 'PC6', fill = 'Sex', point_size = 2)
scatter(x = 'PC5', y = 'PC6', fill = 'Row_numeric', point_size = 2)
scatter(x = 'PC7', y = 'PC8', fill = 'Sex', point_size = 2)
scatter(x = 'PC7', y = 'PC8', fill = 'Row_numeric', point_size = 2)
scatter(x = 'PC9', y = 'PC10', fill = 'Sex', point_size = 2)

scatter(x = 'PC10', y = 'PC11', fill = 'Prob_SNP_outlier', point_size = 2)
scatter(x = 'PC12', y = 'PC13', fill = 'Prob_SNP_outlier', point_size = 2)

pDat_1 %>%
  arrange(Tissue, cor_to_reference) %>%
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = cor_to_reference, fill = Tissue)) +
  geom_point(shape = 21, col = 'black', size = 2.5) + 
  geom_text_repel(data = pDat_1 %>% filter(Sample_Name == 'PM364_hofb_cs'),
                  aes(x = Sample_Name, y = cor_to_reference, label = Sample_Name),
                  nudge_y = -0.01) +
  labs(x = 'Samples', y ='',  title = '')+
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) 
```

## UMAP

```{r}
custom_settings <- umap.defaults
custom_settings$n_neighbors <- 6
custom_settings$random_state <- 2

umap_res <- umap(t(betas_1), config = custom_settings)

umap_res$config
x <- umap_res$layout %>% as_tibble() %>% dplyr::rename(UMAP1 = V1, UMAP2 = V2) %>% 
  bind_cols(pDat_1 %>% select(Tissue, Sample_Name))

ggplot(x, aes(x = UMAP1, y = UMAP2, fill = Tissue)) +
  geom_point(shape = 21, size = 3) + theme_bw() +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  geom_text_repel(data = x %>% filter(Sample_Name %in% c('PM364_hofb_cs', 'PM364_troph_cs')),
                  aes(x = UMAP1, y = UMAP2, label = Sample_Name), nudge_y = 3.5,
                  force = 15)

```

## Hierarchical clustering

```{r}
# assign sample names as numbers
x <- pDat_1 %>%  select(Sample_Name, Case_ID, Tissue)  %>%
  mutate(Sample_number = as.numeric(as.factor(Case_ID))) %>%
  mutate(Sample_ID = paste0(Sample_number, 
                            '_', 
                            str_extract(pDat_1$Sample_Name, '(?<=_).*')))
x

# Dissimilarity matrix
colnames(betas_1) <- x$Sample_ID

d <- dist(t(betas_1), method = "euclidean")

# Hierarchical clustering using Complete Linkage
hc1 <- hclust(d, method = "complete" )
dend <- hc1 %>% as.dendrogram

# plot
par(mar = c(8,4,7,1))
dend %>% plot()

# add colored bars
col_bars <- pDat_1 %>% select(Colors_Sex, Colors_Tissue)

colored_bars(colors = col_bars, 
             dend = dend, 
             sort_by_labels_order = T)

legend(legend = names(color_code_tissue[unique(pDat_1$Tissue)]), 
       fill = color_code_tissue[unique(pDat_1$Tissue)], 
       'top', ncol = 3,  cex = .75, pt.cex = 3.5, inset = c(-0.2), xpd=TRUE)
```

## Correlations

Pairwise

```{r}
# calculate correlation
colnames(betas_1) <- pDat_1$Sample_Name
cor_betas <- cor(betas_1)

#prepare for plot
colnames(cor_betas) <- rownames(cor_betas) <- x$Sample_ID

# annotation bar
annotation <- pDat_1 %>% select(Tissue, Sex) %>% as.data.frame()
rownames(annotation) <- x$Sample_ID

# colors for annotation
anno_colors <- list(
  Tissue = color_code_tissue[unique(pDat_1$Tissue)],
  Sex = setNames(color_code[[2]]$Colors_Sex, color_code[[2]]$Sex)
)

# pheatmap
pheatmap(cor_betas, annotation_col = annotation, show_rownames = F, show_colnames = F,
         labels_col = x$Sample_number, labels_row = x$Sample_number,
         angle_col = 45, fontsize_row = 6, fontsize_col = 6,
         annotation_colors = anno_colors,
         color = viridis(100, option = 'B'),
         cutree_cols = 6)
```

Within and between tissue

```{r}
# sample names as row/colnames
colnames(cor_betas) <- rownames(cor_betas) <- pDat_1$Sample_Name

# tidy up the correlation data, compute average tissue correlations
cor_tissue <- cor_betas %>% as_tibble() %>% 
  
  # gather
  mutate(Sample1 = rownames(cor_betas)) %>%
  gather(key = 'Sample2', value = 'Cor', -Sample1) %>%
  
  # add tissue data for sample1 and sample2
  left_join(pDat %>% select(Sample_Name, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  dplyr::rename(Tissue1 = Tissue) %>%
  left_join(pDat %>% select(Sample_Name, Tissue), by = c('Sample2' = 'Sample_Name')) %>%
  dplyr::rename(Tissue2 = Tissue) %>%
  
  # compute Tissuespecific correlations
  group_by(Sample1, Tissue2) %>%
  summarize(Average_cor = mean(Cor)) %>% 
  ungroup() %>%
  left_join(pDat %>% select(Sample_Name, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  dplyr::rename(Tissue1 = Tissue) 

p1 <- cor_tissue %>%
  arrange(Tissue1, Tissue2, Average_cor) %>%
  mutate(Sample1 = factor(Sample1, levels = unique(Sample1))) %>% 
  ggplot(aes(x = Sample1, y = Average_cor, fill = Tissue1)) +
  geom_point(shape = 21, color = 'black') +
  facet_wrap(~Tissue2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  scale_x_discrete(labels = NULL) +
  theme_bw() + theme(panel.grid.major.x = element_blank()) +
  labs(x = 'Samples', y = 'Correlation', 
       title = 'Average tissue-specific correlation', fill = '')
p1
p1 +
  geom_text_repel(aes(label = ifelse(Sample1 == 'PM364_hofb_cs', 'PM364_hofb_cs', NA)), na.rm = T)


# save to pData
cor_tissue_spread <- cor_tissue %>% mutate(Tissue2 = case_when(
  Tissue2 == 'Endothelial' ~ 'Ave_cor_endo_cs',
  Tissue2 == 'Hofbauer' ~ 'Ave_cor_hofb_cs',
  Tissue2 == 'Stromal' ~ 'Ave_cor_strom_cs',
  Tissue2 == 'Syncytiotrophoblast' ~ 'Ave_cor_troph_enz',
  Tissue2 == 'Trophoblasts' ~ 'Ave_cor_troph_cs',
  Tissue2 == 'Villi' ~ 'Ave_cor_villi',
)) %>%
  spread(key = Tissue2, value = Average_cor)

pDat_1 <- pDat_1 %>% 
  left_join(cor_tissue_spread %>% select(-Tissue1), by = c('Sample_Name' = 'Sample1'))
```


# SessionInfo

```{r}
sessionInfo()
```

# Save data

```{r}
# all necessary for creating report
saveRDS(pDat_1, '../../data/main/interim/21_pDat_term_cells.rds')
saveRDS(cov_tests, '../../data/main/interim/21_cov_tests.rds')
saveRDS(sample_densities, '../../data/main/interim/21_sample_densities.rds')
saveRDS(list(pc_cor = pc_cor, pc_info = pc_info), '../../data/main/interim/21_pca.rds')
saveRDS(list(cor_betas = cor_betas, annotation = annotation, anno_colors = anno_colors,
             x = x, cor_tissue = cor_tissue), '../../data/main/interim/21_pheatmap_data.rds')

# additional stuff:
saveRDS(d, '../../data/main/interim/21_dist.rds') # distance matrix
```
