---
title: "Removing contaminated term samples"
output:
  html_document:
    keep_md: false
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
    self_contained: yes
editor_options: 
  chunk_output_type: console
---

May 31st, 2018 update:
Originally this report was to document the reasons for removal of PM364_hofb. However, I am now
removing 2 additional hofbauer samples, so I have added some additional analysis at the end of this report explaining the removal of these samples. Everything up to that point is unedited.

# May 31st Summary
* remove PM364_hofb, because contaminated with HOFBAEUR cells
* remove PM381_hofb, PM366_hofb because contaminated with maternal cells
* PM374_endo_cs, may be contaminated with endothelial cells from another sample, retaining.

PM364_hofb is an 'odd' sample compared to other third trimester cell samples. So I'm wondering if 
this sample should be removed.

# PM364_hofb_cs
*begin original report*

**Main observations:**

* unusual density distribution on cpg betas methylation
* clusters with trophoblast cells

*This is unlikely a quality issue, because:*

* QC thresholds were met
* CNV plot looks clean/tight (not noisy)

*It is unlikely from contamination with maternal cells:*

* SNP correlation to reference villi is > 0.98

These observations make me believe it is contaminated with trophoblast cells, which I think warrants
removal, but what do you think?

```{r include = FALSE}
knitr::opts_chunk$set(echo=FALSE) #hide code
```

```{r setup, message = F, warning = F, include = F}
# libraries and data
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggrepel)
library(viridis)
library(scales)
library(RColorBrewer)
library(pheatmap)
library(GGally)
library(cowplot)
library(tibble)
library(kableExtra)
library(egg)

pDat <- readRDS('../../data/main/interim/13_pDat.rds')
pDat_1 <- readRDS('../../data/main/interim/21_pDat_term_cells.rds')
cov_tests <- readRDS('../../data/main/interim/21_cov_tests.rds')
sample_densities <- readRDS('../../data/main/interim/21_sample_densities.rds')
pca <- readRDS('../../data/main/interim/21_pca.rds')
pheatmap_data <- readRDS('../../data/main/interim/21_pheatmap_data.rds')
color_code <- readRDS('../../data/main/interim/11_color_code.rds')
color_code_tissue <- setNames(color_code[[1]]$Colors_Tissue, color_code[[1]]$Tissue)
```

# Setup

Below analyses were conducted on 103 third trimester samples from the following cell/tissue types:

```{r}
pDat_1 %>% dplyr::count(Tissue) %>% kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, position = 'left')
```

The following were excluded for my preliminary analysis, but will be included in later ones:

```{r, message = F}
pDat %>% 
  anti_join(pDat_1 %>% select(Sample_Name)) %>%
  filter(Trimester == 'Third') %>% 
  dplyr::count(Tissue) %>% kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, position = 'left')
```

Note the two villi are PM324 V1 and V4 samples, ran for follow-up on GDG's work.

# Sample density plots

PM364_hofb appears to have a large peak at 0.5-0.6, similar to what is seen in the trophoblast 
cells.

```{r}
ggplot(sample_densities, aes(x = x, y = y, color = Tissue, group = Sample_Name)) +
  geom_line(size = 1, alpha = 0.5) + theme_bw() +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1))  +
  facet_wrap(~Tissue) +
  scale_color_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  geom_text_repel(data = sample_densities %>% 
                    filter(Sample_Name == 'PM364_hofb_cs', x > 0.55, x < 0.6) %>%
                    dplyr::slice(1),
                  aes(x = x, y = y, label = Sample_Name), inherit.aes = F,
                  nudge_y = 1, nudge_x = -0.1) +
  labs(x = '% methylation', y = 'density')
```

# PCA

We see here that PC1, PC2, and PC3 are associated with Tissue.

```{r}
colpal <- c('white', '#fee8c8', '#fdbb84', '#e34a33')
names(colpal) <- levels(pca$pc_cor$pval_cat)

p1 <- ggplot(pca$pc_cor, aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile(col = 'lightgrey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0), labels = 1:20) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = colpal)  + 
  labs(y = '', fill = 'P value')

p2 <- ggplot(pca$pc_info %>% filter(variable == 'Proportion of Variance') %>%
               mutate(value = value*100), 
             aes(x = PC, y = value)) +
  geom_bar(stat = 'identity') +
  theme_bw() + 
  scale_x_discrete(expand = c(0, 0), labels = 1:20) + 
  scale_y_continuous(expand = c(0, 0)) +
  labs(y = '% variance explained')

egg::ggarrange(p1, p2, heights = c(3,1))
```

## PC scatter plots { .tabset}

```{r}
scatter <- function(x, y, fill, point_size = 1){
  xlab <- pca$pc_info %>% filter(variable == 'Proportion of Variance', PC == x) %>% pull(Label)
  ylab <- pca$pc_info %>% filter(variable == 'Proportion of Variance', PC == y) %>% pull(Label)
  
  x <- paste0(x, '_processed')
  y <- paste0(y, '_processed')
  
  out <- ggplot(pDat_1, aes_string(x = x, y = y, fill = fill)) +
    geom_point(shape = 21, size = point_size) + theme_bw() + labs(x = xlab, y = ylab) 
  
  if (is.numeric(as.data.frame(pDat_1)[,fill])){
    out <- out +
      scale_fill_viridis()
  } else {
    out <- out + 
      scale_fill_brewer(palette = 'Set1')
  }
    
  out
}
```

### PC1 & PC2

PM364_hofb_cs appears to cluster with trophoblast/villi samples rather than hofbaeur cells. This is
unlikely due to a sample swap with PM364_troph, since PM364_troph clusters normally with other 
trophoblast cells.

Desmond also mentioned previously that cross-contamination during sorting was a concern, however 
looking through his notes, I haven't seen anything noted on this particular sort.

```{r, message = F}
scatter(x = 'PC1', y = 'PC2', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)])
scatter(x = 'PC1', y = 'PC2', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  geom_text_repel(data = pDat_1 %>% filter((Tissue == 'Hofbauer' & PC1_processed < 0)|
                                             Case_ID == 'PM364'),
                  aes(x = PC1_processed, y = PC2_processed, label =Sample_Name),
                  nudge_y = 15, nudge_x = -5)
```

### PC3 & PC4

```{r , message = F}
scatter(x = 'PC3', y = 'PC4', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)])
scatter(x = 'PC3', y = 'PC4', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  geom_text_repel(data = pDat_1 %>% filter(Case_ID == 'PM364'),
                  aes(x = PC3_processed, y = PC4_processed, label =Sample_Name),
                  force = 50)
```

It might be possible that this is also due to maternal contamination. However, PM364_hofb has very
low levels of contamination:

```{r}
pDat_1 %>%
  arrange(Tissue, cor_to_reference) %>%
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = cor_to_reference, fill = Tissue)) +
  geom_point(shape = 21, col = 'black', size = 2.5) + 
  geom_text_repel(data = pDat_1 %>% filter(Sample_Name == 'PM364_hofb_cs' | cor_to_reference < 0.96),
                  aes(x = Sample_Name, y = cor_to_reference, label = Sample_Name),
                  nudge_y = -0.01, force = 10) +
  labs(x = 'Samples', y ='Correlation to reference villi',  title = '')+
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) 
```

We can see that other hofbauer cells have some level of contamination, but do not cluster with 
PM364_hofb.

# Correlation

Pairwise correlations on the processed DNAme data also show PM364_hofb clusters with 
villi/trophoblast samples:

```{r}
rownames(pheatmap_data$annotation) <- pDat_1$Sample_Name

pheatmap_data$annotation$PM374_endo <- ifelse(rownames(pheatmap_data$annotation) == 'PM374_endo_cs',
                                              'T',
                                              'F')
pheatmap_data$anno_colors$PM374_endo <- setNames(c('Red', 'Grey'), c('T', 'F'))

# pheatmap
pheatmap(pheatmap_data$cor_betas, annotation_col = pheatmap_data$annotation, 
         annotation_row = pheatmap_data$annotation[,c('Tissue'), drop = F],
         show_rownames = F, show_colnames = F,
         labels_col = pheatmap_data$x$Sample_number, labels_row = pheatmap_data$x$Sample_number,
         angle_col = 45, fontsize_row = 6, fontsize_col = 6,
         annotation_colors = pheatmap_data$anno_colors,
         color = viridis(100, option = 'B'),
         cutree_cols = 6)
```

Then we can average each samples average correlation to all samples on a per-tissue basis.

```{r}
p1 <- pheatmap_data$cor_tissue %>%
  arrange(Tissue1, Tissue2, Average_cor) %>%
  mutate(Sample1 = factor(Sample1, levels = unique(Sample1))) %>% 
  ggplot(aes(x = Sample1, y = Average_cor, fill = Tissue1)) +
  geom_point(shape = 21, color = 'black') +
  facet_wrap(~Tissue2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  scale_x_discrete(labels = NULL) +
  theme_bw() + theme(panel.grid.major.x = element_blank()) +
  labs(x = 'Samples', y = 'Correlation', 
       title = 'Average tissue-specific correlation', fill = '')
p1
```

# Maternal -contaminated samples

At this point, I have decided to remove PM364_hofb_cs. 

We previously noticed a few term sorted samples had maternal contamination. So here I return to this
analysis and decide on which samples to remove. 

```{r}
p3 <- pDat_1 %>%
  arrange(Tissue, cor_to_reference) %>%
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = cor_to_reference, fill = Tissue)) +
  geom_point(shape = 21, col = 'black', size = 2.5) + 
  geom_text_repel(data = pDat_1 %>% filter(Sample_Name == 'PM364_hofb_cs'| 
                                             cor_to_reference < 0.96),
                  aes(x = Sample_Name, y = cor_to_reference, label = Sample_Name),
                  nudge_y = -0.01, force = 15) +
  labs(x = 'Samples', y =expression(rho),  title = 'Correlation to villi', color = '')+
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) 

p4 <- pDat_1 %>%
  arrange(Tissue, Prob_SNP_outlier) %>%
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = Prob_SNP_outlier, fill = Tissue)) +
  geom_point(shape = 21, col = 'black', size = 2.5) + 
  geom_text_repel(data = pDat_1 %>% filter(Sample_Name == 'PM364_hofb_cs'| 
                                             Sample_Name == 'PM374_endo_cs' |
                                             Prob_SNP_outlier > 0.3),
                  aes(x = Sample_Name, y = Prob_SNP_outlier, label = Sample_Name),
                  nudge_y = -0.01) +
  labs(x = 'Samples', y ='P',  title = 'Probability SNP is an outlier', color = '')+
  scale_fill_manual(values = color_code_tissue[unique(pDat_1$Tissue)]) +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) 

ggarrange(p3, p4)
```

Samples PM366_hofb, PM381_hofb, and PM374_endo have a low correlation to villi. However, only PM381_hofb, PM366_hofb have a high probability of being an outlier. This suggests that either PM374_endo is mainly maternal endothelial cells or is contaminated with another sample. 


Becayse PM374_endo is a male sample we can confirmt his based on XY intensity:

```{r}
contam_samps <- c('PM364_hofb_cs', 'PM374_endo_cs', 'PM381_hofb_cs', 'PM366_hofb_cs')

p5 <-ggplot(pDat_1, aes(x = normalized_X_intensity, y = normalized_Y_intensity, fill = Sex)) +
  geom_point(shape = 21, col = 'black', size = 2.5) +theme_bw() +
  geom_text_repel(data = pDat %>% filter(Sample_Name %in% contam_samps),
                  aes(label = Sample_Name), size = 3, force = 15, nudge_x = -0.1, nudge_y = -0.1) +
  geom_hline(yintercept = 0.5, linetype = 'dashed', col = '#9B9B9B') + 
  geom_vline(xintercept = 0.95, linetype = 'dashed', col = '#9B9B9B')
p5
```

Surprisingly, PM374_endo looks 100% male on this plot, suggesting that it is not contaminated with 
maternal cells. The low correlation to villi then suggests it is contaminated. Another possibility 
is it is poor quality, but it was not flagged in previous analyses. We leave it in for now...

So in summary I remove the following samples:

* remove PM364_hofb, because contaminated with HOFBAEUR cells
* remove PM381_hofb, PM366_hofb because contaminated with maternal cells
* PM374_endo_cs, may be contaminated with endothelial cells from another sample, retaining.
