---
title: "1_3_maternal_contamination"
author: "Victor Yuan"
date: "May 21, 2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

# Setup

```{r, warning = F, message = F}
library(minfi)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ewastools)
library(ggrepel)
library(viridis)
library(GGally) # for ggcorr
source('../check_snp_agreement2.R') # code to return all agreement scores

pDat <- readRDS('../../data/main/interim/1_1_pDat.rds')

# called genotypes
snps_called <- readRDS('../../data/main/interim/1_1_snps_called.rds')


cor_snp <- readRDS('../../data/main/interim/1_1_cor_matrix_snp_betas.rds')
```

# Limitations of inferring contamination

Normalized Y intensity only applicable for male samples
Probability snp outlier only linearly correlated with the minority genotype

```{r}
ggplot(pDat, aes(x = normalized_X_intensity, y = normalized_Y_intensity, 
                 fill = Prob_SNP_outlier)) +
  geom_point(aes(shape = Sex), size = 2.5, alpha = 0.95) +
  scale_shape_manual(values = c('M' = 21, 'F' = 22)) +
  theme_bw() +
  scale_fill_viridis_c(option = 'inferno')

ggplot(pDat %>% filter(Sex == 'M'), 
       aes(x = normalized_Y_intensity, y = Prob_SNP_outlier, fill = Prob_SNP_outlier)) +
  geom_point(shape = 21, color = 'black', size = 3) + 
  geom_smooth(method = 'loess')+
  theme_bw() +
  scale_fill_viridis_c(option = 'inferno') 

ggplot(pDat %>% filter(Sex == 'M'), 
       aes(x = normalized_Y_intensity, y = Prob_SNP_outlier)) +
  geom_smooth(method = 'loess')+
  geom_point(shape = 21, color = 'black', size = 3, aes(fill = Tissue)) + 
  theme_bw() +
  scale_fill_brewer(palette = 'Set1')
```

# Agreement between inferred snps

**Motivation**: Based on previous data, villi are relatively clean (see snp distributions in QC report). Therefore, we can safely assume that fetal genotype is represented by the villi samples. Deviations from villi genotypes, therefore should be a proxy for maternal contamination.

**The measure of maternal contamination:** Based on the probability each sample for each snp is belonging to one of three genotypes, we can determine the proportion of SNPs that agree between all pairwise comparisons between samples. Grouping comparison to those between a villi sample and other samples from the same placenta, we can gain a measure of maternal contamination (assuming that deviation from fetal genotype is always maternal).

To do this, I turn to two approaches. #1 involves using the output of ewastools::check_snp_agreement,
and #2 involves correlations on the snp betas.

## ewastools

I edited the `check_snp_agreement` function and saved as `check_snp_agreement2`, so that all comparisons are returned.

```{r}
# original function
check_snp_agreement(snps_called, donor_ids = pDat$Case_ID, sample_ids = pDat$Sample_Name) 
  
# percent of SNPs concordant 
agreement <- check_snp_agreement2(snps_called, threshold = F, return = 'within-donor',
                                  donor_ids = pDat$Case_ID, sample_ids = pDat$Sample_Name) 

# Filter to within donor group comparisons
agreement <- agreement %>% 
  left_join(pDat %>% select(Sample_Name, Tissue, Prob_SNP_outlier), 
            by = c('sample2' = 'Sample_Name')) %>%
  rename(Tissue2 = Tissue, P_outlier2 = Prob_SNP_outlier) %>%
  left_join(pDat %>% select(Sample_Name, Tissue, Prob_SNP_outlier), 
            by = c('sample1' = 'Sample_Name')) %>%
  rename(Tissue1 = Tissue, P_outlier1 = Prob_SNP_outlier) %>%
  filter(Tissue1 == 'Villi' | Tissue2 == 'Villi') %>%
  select(contains('1'), contains('2'), everything()) %>% as_tibble %>%
  arrange(donor1, donor2, sample1, sample2)

# swap columns such that villi is always in 'sample1'
agreement <- agreement %>% 
  mutate(sample_reference = case_when(Tissue1 == 'Villi' ~ sample1,
                                      Tissue2 == 'Villi' ~ sample2), 
         sample_query = case_when(Tissue1 == 'Villi' ~ sample2,
                                  Tissue2 == 'Villi' ~ sample1),
         
         tissue_query = case_when(Tissue1 == 'Villi' ~ Tissue2,
                                  Tissue2 == 'Villi' ~ Tissue1)) %>%
  rename(donor = donor1) %>%
  select(sample_reference, sample_query, tissue_query, 
         donor, agreement, contains('outlier')) %>%
  arrange(donor, tissue_query) %>%

# average agreement scores when comparing to replicate villi samples
  group_by(donor, tissue_query) %>%
  mutate(agreement_ave = mean(agreement)) %>% ungroup() %>%
  select(sample_query, agreement_ave) %>%
  distinct()

# add to pData
pDat <- pDat %>% 
  left_join(agreement, by = c('Sample_Name' = 'sample_query')) %>%
  rename(Agreement_to_donor_villi = agreement_ave) 
```

For a reference I compare to the average agreement of samples to non-donor samples

```{r}
# return comparisons for ALL samples
agreement <- check_snp_agreement2(snps_called, threshold = F, return = 'all',
                                  donor_ids = pDat$Case_ID, sample_ids = pDat$Sample_Name) 

# filter to out of donor sample  comparisons 
agreement <- agreement %>% as_tibble() %>%
  filter(donor1 != donor2) %>%
  
  # duplicate the rows so that all comparisons are represented in sample1 column
  bind_rows(agreement %>% rename(donor1 = donor2, sample1 = sample2,
                                 donor2 = donor1, sample2 = sample1)) %>%
  arrange(sample1, sample2)

# check that around 180-190 comparisons for each sample 
table(agreement$sample1)

# calculate average out of donor SNP agreement scores
agreement <- agreement %>% group_by(sample1) %>%
  summarize(Agreement_to_unrelated = mean(agreement))

pDat <- pDat %>% left_join(agreement, by = c('Sample_Name' = 'sample1')) 

# plot agremeent scores
pDat %>% arrange(Tissue, desc(Agreement_to_donor_villi)) %>% 
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = Agreement_to_donor_villi, color = Tissue)) +
  geom_point() +
  scale_color_brewer(palette = 'Set1') +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1, 0.1)) +
  geom_hline(yintercept = mean(pDat$Agreement_to_unrelated), linetype = 'dashed', color = 'red')  +
  facet_grid(~Trimester, scales = 'free_x', space = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank())
```

How does this compare to previous measures of contamination?

```{r}
ggplot(pDat, aes(x = Agreement_to_donor_villi, y = normalized_Y_intensity, color = Sex)) +
  geom_point() + theme_bw() +
  geom_smooth(method = 'lm')
ggplot(pDat, aes(x = Agreement_to_donor_villi, y = Prob_SNP_outlier, color = Sex)) +
  geom_point() + theme_bw()
```

In males, SNP agreement to villi is highly correlated with normalized Y intensity.

probability of having SNPs being outliers shows the mountain shaped pattern described previously.

The limitation of this approach, however, is that this requires a clean reference.

```{r}
# samples without sex information
ggplot(pDat, aes(x = normalized_X_intensity, y= normalized_Y_intensity, color = Sex)) +
  geom_point() + theme_bw() +
  geom_text_repel(data = pDat %>% filter(Case_ID %in% c('P131', 'P128')),
             aes(x = normalized_X_intensity, y= normalized_Y_intensity, label = Case_ID),
             color = 'black', nudge_x = -0.15, nudge_y = 0.1, force = 25)+
  geom_point(data = pDat %>% filter(Case_ID %in% c('P131', 'P128')),
             aes(x = normalized_X_intensity, y= normalized_Y_intensity),
             color = 'black')
```

## Correlation to donor villi

The idea here is the same, except instead of posterior probabilities on the snp probes, we use the
pairwise correlations on the snp betas. 

```{r}
# correlation to donor villi
cor_to_donor_villi <- cor_snp %>% as_tibble() %>% mutate(Sample1 = rownames(cor_snp)) %>% 
  left_join(pDat %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  tidyr::gather(key = 'Sample2', value = 'Cor', -Sample1, -Case_ID, -Tissue) %>%
  dplyr::rename(Case_ID1 = Case_ID, Tissue1 = Tissue) %>%
  left_join(pDat %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample2' = 'Sample_Name')) %>%
  dplyr::rename(Case_ID2 = Case_ID, Tissue2 = Tissue) %>% 
  filter(Case_ID1 == Case_ID2, Tissue2 == 'Villi') %>%
  group_by(Sample1) %>%
  summarize(cor_to_donor_villi = mean(Cor))

pDat <- pDat %>% left_join(cor_to_donor_villi, by = c('Sample_Name' = 'Sample1'))

# correlation to unrelated samples
cor_to_unrelated <- cor_snp %>% as_tibble() %>% mutate(Sample1 = rownames(cor_snp)) %>% 
  left_join(pDat %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  tidyr::gather(key = 'Sample2', value = 'Cor', -Sample1, -Case_ID, -Tissue) %>%
  dplyr::rename(Case_ID1 = Case_ID, Tissue1 = Tissue) %>%
  left_join(pDat %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample2' = 'Sample_Name')) %>%
  dplyr::rename(Case_ID2 = Case_ID, Tissue2 = Tissue) %>% 
  filter(Case_ID1 != Case_ID2) %>%
  group_by(Sample1) %>%
  summarize(cor_to_unrelated = mean(Cor))

pDat <- pDat %>% left_join(cor_to_unrelated, by = c('Sample_Name' = 'Sample1'))

# correlation to donor villi compared to unrelated
pDat %>% select(cor_to_donor_villi, cor_to_unrelated) %>% gather() %>%
  ggplot(aes(x = key, y = value)) + theme_bw() + geom_boxplot()

# correlation to other measures of contamination
ggplot(pDat, aes(x = cor_to_donor_villi, y = Agreement_to_donor_villi)) +
  geom_point() + theme_bw() +stat_smooth(method = 'lm')

fit <- lm(cor_to_donor_villi ~ normalized_Y_intensity, data = pDat %>% filter(Sex == 'M'))
fit <- summary(fit)
ggplot(pDat, aes(x = cor_to_donor_villi, y = normalized_Y_intensity, color = Sex)) + 
  geom_point() + theme_bw()+stat_smooth(method = 'lm') +
  geom_text(x = 0.7, y = 0.9, color = 'black',
            label = paste0('Rsquared = ', prettyNum(fit$r.squared, digits = 3), 
                           '\np-value = ', prettyNum(fit$coefficients[2,4], digits = 3)))


pDat %>% arrange(Tissue, desc(cor_to_donor_villi)) %>% 
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = cor_to_donor_villi, color = Tissue)) +
  geom_point() +
  scale_color_brewer(palette = 'Set1') +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1, 0.1)) +
  facet_grid(~Trimester, scales = 'free_x', space = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) +
  geom_hline(yintercept = mean(pDat$cor_to_unrelated), linetype = 'dashed', color = 'red')
```

Because we are missing villi samples for two first trimester cases, we are also missing inferred
contamination information. However, if we inspect the pairwise correlations, we can see it's likely
OK to pick a different cell type to serve as reference for fetal genotype. Also, based on the other
samples, we can see that first trimester stromal and endothelial cells are less often contaminated.

```{r}
ind <- pDat %>% 
  add_count(Case_ID) %>%
  filter(Case_ID %in% c('P128', 'P131'))  %>%
  arrange(n, Case_ID) %>%
  pull(Sample_Name)

# plot settings
label_size <- 5
point_size <- 13
r_size <- 4
colors <- c('red', 'red', rev(viridis_pal(option = 'D')(9))[c(1,1,1:8)])
r_limit <- c(0.65,1)

ggcorr(data = NULL, cor_matrix = cor_snp[ind,ind], 
             geom = 'blank', name = expression(rho), size = label_size, hjust = 0.95) +
  geom_point(aes(col = coefficient), size = point_size) +
  geom_text(data = function(x) dplyr::filter(x, coefficient > r_limit[1]), 
            aes(label = prettyNum(coefficient, digits = 2)),
            size = r_size) +
  scale_color_gradientn(colours = colors, 
                        limits = r_limit, na.value = 'white')

ggplot(pDat %>% filter(Sample_Name %in% ind), aes(x = Sample_Name, y = Prob_SNP_outlier)) +
  geom_bar(stat = 'identity')

# check to agreement scores
agreement <- check_snp_agreement2(snps_called, threshold = F, return = 'all',
                                  donor_ids = pDat$Case_ID, sample_ids = pDat$Sample_Name) 
agreement %>% filter(donor1 %in% c('P128', 'P131') | donor2 %in% c('P128', '131'), donor1 == donor2)
```

Below I recompute correlations for all samples, making sure that for these two first trimester 
samples, I use stromal cells to serve as the reference genotype (Whereas villi is used for all 
other samples).

```{r}
cor_to_reference <- cor_snp %>% as_tibble() %>% mutate(Sample1 = rownames(cor_snp)) %>% 
  left_join(pDat %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  tidyr::gather(key = 'Sample2', value = 'Cor', -Sample1, -Case_ID, -Tissue) %>%
  dplyr::rename(Case_ID1 = Case_ID, Tissue1 = Tissue) %>%
  left_join(pDat %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample2' = 'Sample_Name')) %>%
  dplyr::rename(Case_ID2 = Case_ID, Tissue2 = Tissue) %>% 
  filter(Case_ID1 == Case_ID2, 
         Tissue2 == 'Villi' | (Case_ID2 %in% c('P128', 'P131') & Tissue2 == 'Stromal')) %>%
  group_by(Sample1) %>%
  summarize(cor_to_reference = mean(Cor))

pDat <- pDat %>% left_join(cor_to_reference, by = c('Sample_Name' = 'Sample1'))


pDat %>% arrange(Tissue, desc(cor_to_reference)) %>% 
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = cor_to_reference, color = Tissue)) +
  geom_point() +
  scale_color_brewer(palette = 'Set1') +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1, 0.1)) +
  facet_grid(~Trimester, scales = 'free_x', space = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) +
  geom_hline(yintercept = mean(pDat$cor_to_unrelated), linetype = 'dashed', color = 'red')
```

# SessionINfo

```{r}
sessionInfo()
```

# save data
```{r, eval = F}
pDat <- pDat %>% 
  select(Sample_Name:Prob_SNP_outlier_Logodds, 
         Agreement_to_donor_villi:cor_to_reference,
         PC1_raw:PC20_raw)
saveRDS(pDat, '../../data/main/interim/1_3_pDat.rds')
```