---
title: "Technical variables in EPIC array analysis"
output:
  html_document:
    keep_md: false
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
    self_contained: yes
editor_options: 
  chunk_output_type: console
---

Here I examine some of the technical factors that can affect EPIC data


```{r include = FALSE}
knitr::opts_chunk$set(echo=FALSE) #hide code
```

```{r setup, message = F, warning = F, include = F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggrepel)
library(viridis)
library(yahew)
library(egg)

pDat <- readRDS('../../data/main/interim/13_pDat.rds')
# PCA association data for plotting
pca <- readRDS('../../data/main/interim/11_pca_association_plot.rds')
```

I inspected the association between various sample characteristics for all 192 samples:

```{r warning = F}
cov_tests <- pDat %>%
  select(Row_numeric, Row_factor, Chip_number, Week,Batch_BSC,
         DNA_conc_before_load, DNA_loaded,  
         failed_probes, detP_01_minfi , beadcount, Average_intensity, 
         Case_ID, Sex, Tissue,  Trimester,
         cor_to_reference, cor_to_unrelated, Prob_SNP_outlier) %>%
  as.data.frame() %>% pairtest

# make categories
cov_tests <- cov_tests %>% 
  mutate(pval_cat = if_else(p.value < 0.001, '< 0.001',
                            if_else(p.value < 0.01, '< 0.01',
                                    if_else(p.value < 0.05, '< 0.05', '> 0.05'))))

# plot heatmap of associations
ggplot(cov_tests, aes(x=Column, y = Row, fill = pval_cat)) +
  geom_tile(col = 'grey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = c('> 0.05' = 'White', '< 0.05' = '#fee8c8', 
                               '< 0.01' = '#fdbb84', '< 0.001' = '#e34a33')) +
  labs(x = '', y = '', fill = 'p') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust= 1),
        panel.grid.major.x = element_blank()) +
  coord_equal()
```

After plotting and going through each significant association manually, I found some to be driven
by outliers and are probably not actually real, and a few that were convincing. Some of these might
be good to know for future runs.

# DNA loaded and number of failed probes

There is a weak association between the amount of DNA loaded onto the array and failed probes. The number of failed probes increases only after decreasing from DNA loaded < 100 ng. In fact, the majority of the samples with a high number of failed probes  had > 150ng loaded.

```{r warning = F, message = F}
qplot(data = pDat,x = DNA_loaded, y = failed_probes) + geom_smooth()
```

# Row

Row and failed probes were not strongly associated 

```{r, warning = F, message = F}
qplot(data = pDat, x = Row_numeric, y = failed_probes) + geom_smooth()
```

But Row is associated with the number of probes failing detection p value threshold. The number
of failed probes increases starting at 5 until row 8. But in the first 4 rows, there is basically
no change.

```{r, warning = F, message = F}
qplot(data = pDat, x = Row_numeric, y = detP_01_minfi) + geom_smooth()
```

Average intensity is highly assocaited with Row. Again we see an affect in only a subset of the rows.
Average intensity increases from Row 1 -> Row 5, and then plateaus.

```{r, warning = F, message = F}
qplot(data = pDat, x = Row_numeric, y = Average_intensity) + geom_smooth()
```

# BSC Batch

We see a small association of average intensity with BSC Batch, where BSC0 had a lower intensity 
compared to all other batches. Although this is confounded with the fact that BSC0 only had mixtures,
which were generally a bit lower in the DNA loaded.

```{r, message = F}
qplot(data = pDat, x = as.numeric(as.factor(Batch_BSC)), y =Average_intensity) + 
  geom_smooth(method = 'lm') + 
  scale_x_continuous(labels = levels(as.factor(pDat$Batch_BSC))) +
  scale_y_continuous(limits = c(0, 14500))
```

# PCA {.tabset}

I previously showed no row effects. I originally had row encoded as a "factor" variable, but now I 
have changed it to both numeric and factor. Now it appears associated with a number of PCs.

I also found that my DNA concentrations were encoded as 0, when really they should have been NAs, 
as 0 means we did not measure it for that sample. But this did not result in any significant changes.

```{r, message = F}
colpal <- c('white', '#fee8c8', '#fdbb84', '#e34a33')
names(colpal) <- levels(pca$Plot_data$pval_cat)

p1 <- ggplot(pca$Plot_data, aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile(col = 'lightgrey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0), labels = 1:20) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = colpal)  + 
  labs(y = '', fill = 'P value')

p1b <- ggplot(pca$PC_variance, aes(x = PC, y = Prop_var_raw)) +
  geom_bar(stat = 'identity') + theme_bw() + labs(y = '% variance') +
  scale_x_continuous(breaks = 1:20)

ggarrange(p1, p1b, heights = c(3,1))
```

After inspecting the PC plots, we can see that the row associations are relatively weak. Their R
squareds are all < 0.15. 

## PC5

```{r, message = F}
qplot(data = pDat, x = Row_numeric, y = PC5_raw) +theme_bw() + geom_smooth()
```

## PC6

```{r, message = F}
qplot(data = pDat, x = Row_numeric, y = PC6_raw) +theme_bw() + geom_smooth()
```

## PC7

```{r, message = F}
qplot(data = pDat, x = Row_numeric, y = PC7_raw) +theme_bw() + geom_smooth()
```

## PC8

```{r, message = F}
qplot(data = pDat, x = Row_numeric, y = PC8_raw) +theme_bw() + geom_smooth()
```

## PC10

```{r, message = F}
qplot(data = pDat, x = Row_numeric, y = PC10_raw) +theme_bw() + geom_smooth()
```