---
title: "12_CNVs"
author: "Victor Yuan"
date: "May 13, 2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

For calling CNVs

# Setup

## Libraries and Data

```{r}
library(minfi)
library(cnAnalysis450k)
library(ggplot2)
library(dplyr)
library(GenVisR)
library(tidyr)

rgset <- readRDS('../../data/main/interim/01_rgset_raw.rds')
pDat <- readRDS('../../data/main/interim/11_pDat.rds')
probe_anno <- readRDS('../../data/main/interim/11_probe_anno.rds')

all(pDat$Sentrix == colnames(rgset)) # T
```

## Normalize

We use noob normalization 

```{r}
mset_noob <- preprocessNoob(rgset)

#get copy number data
cn_noob <- getCN(mset_noob)
colnames(cn_noob) <- pDat$Sample_Name
cn_noob[1:6,1:6]
```

# Conumee

## Setup

Conumee produces an annotation that does not match the probes of current EPIC arrays. Be sure to
run subsetbyoverlaps with the data after running CNV.load

```{r}
# define controls
CNV_controls <- pDat %>% 
  filter(Trimester == 'Third' & Sex == 'M') %>% 
  pull(Sample_Name) %>% as.character()

# exclude cross hyb probes
# load annotation
zhou_anno <- readRDS('Z:/Victor/Data/DNAm annotations/zhou2017_EPIC.hg19.manifest.rds') 

# subset zhou anno to filtered probes
zhou_anno <- zhou_anno[intersect(rownames(mset_noob), names(zhou_anno)),]

exclude <- zhou_anno[zhou_anno$MASK_general == T,]
exclude <- exclude[!seqnames(exclude) %in% c('chrM','*'),]

# create annotation
CNV_anno <- CNV.create_anno(array_type="EPIC", chrXY=TRUE, 
                            exclude_regions = granges(exclude))

# subset annotation to those present in mset
CNV_anno@probes <- subsetByOverlaps(CNV_anno@probes, granges(zhou_anno))

# load into cnv calling format
CNV <- CNV.load(mset_noob[intersect(rownames(mset_noob), names(zhou_anno)),])

#test
#x <- CNV.fit(CNV[2], ref = CNV[setdiff(CNV_controls, names(CNV[2]))], anno = CNV_anno)
#x <- CNV.bin(x)
#x <- CNV.detail(x)
#x <- CNV.segment(x)
#CNV.genomeplot(x)
#CNV.genomeplot(x, chr = 'chr6')
#y2 <- CNV.write(x, what = 'bins')
#head(y2)
```

## Call CNVs

```{r}
# initilize cluster
cl <- makeCluster(24)
registerDoParallel(cl)

# parallel cnv fit
CNV_calls <- foreach (i =  1:ncol(mset_raw),
                          .export = c("CNV.fit", "CNV.bin", 
                                      "CNV.segment", "CNV.detail")) %dopar%
  CNV.segment(
    CNV.detail(
      CNV.bin(
        CNV.fit(
          CNV[i], ref = CNV[setdiff(CNV_controls, names(CNV[i]))], anno = CNV_anno))))

# kill cluster
stopCluster(cl)

# name the output list
names(CNV_calls) <- pDat$Sample_Name
CNV.genomeplot(CNV_calls[[1]])

# visualize 
#organize by case
Cases <- levels(as.factor(as.character(pDat$Case_ID)))
```

## Output genomeplots

```{r eval = F}
pdf(file='../../outs/11_CNV_conumee.pdf')  
for (i in 1:length(Cases)) {
  Case_i <- Cases[i]
  pDat_i <- pDat %>% filter(Case_ID == Case_i)
  name_i <- as.character(pDat_i$Sample_Name)
  sex_i <- as.character(pDat_i$Sex)
  
  
  # set number of plots per page
  if (length(name_i) == 1) {
    par(mfrow = c(1,1))
  }
  
  if (length(name_i) == 2) {
    par(mfrow = c(2, 1)) # two plots on two rows, better than two plots in two columns
  }
  
  if (length(name_i) > 2) {
    par(mfrow = c(ceiling(length(name_i)/2), 2)) # set two columns and enough rows to hold all plots
  }
  
  
  for (j in 1:length(name_i)){
    CNV.genomeplot(CNV_calls[[name_i[j]]],
                   main = paste0(name_i[j], ' (', sex_i[j], ')' ), set_par = F)
  }
}
dev.off()
```

## Output data matrix

probes
bins
segments

```{r}
#segments
CNV_segments <- CNV.write(CNV_calls[[1]], what = 'segments')
CNV_segments <- lapply(CNV_calls, function(x) CNV.write(x, what = 'segments')) %>% bind_rows
head(CNV_segments)

#bins
CNV_bins <- lapply(CNV_calls, function(x) CNV.write(x, what = 'bins')[,5])
names(CNV_bins) <- names(CNV_calls)
CNV_bins <- bind_cols(CNV.write(CNV_calls[[1]], what = 'bins')[,1:4],
                      CNV_bins) %>% as_tibble()
CNV_bins 

```


## Visualize

Need to format so that I can plot as heatmap

Y axis will be position in genome

I need a continuous position along the entire genome, not just the chromosome.

So chromosome 2 positions need to have the maximum chromosome 1 length to them, etc.

```{r}
# create plotting data
plot_data <- CNV_bins %>% 
  mutate(Chromosome = factor(as.character(Chromosome), levels = 
                               paste0('chr', c(1:22, 'X', 'Y'))))

# calculate chromosome lengths
chrom_lengths <- plot_data %>% group_by(Chromosome) %>% 
  dplyr::summarize(max = max(End)) %>%
  mutate(toadd = cumsum(as.numeric(max)) - max) # subtract current chromosome length

plot_data <- chrom_lengths %>% inner_join(plot_data) %>% mutate(position_genome = Start + toadd) %>%
  select(Chromosome, Start, End, Feature, position_genome, everything())

plot_data_melt <- plot_data %>%
  select(-Start, -End, -Feature, -toadd) %>%
  tidyr::gather(key = 'Sample', value = 'Intensity', -Chromosome, -position_genome, -max) %>%
  mutate(State = factor(case_when(
    Intensity < -0.2 ~ 'Loss',
    Intensity > 0.2 ~ 'Gain',
    TRUE ~ 'Neutral'
  ), levels = c('Loss', 'Gain', 'Neutral')))


ggplot(plot_data_melt[1:900000,], 
       aes(x = Sample, y = position_genome, color = State)) +
  geom_point(shape = 15, size = 0.75) +
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(breaks = chrom_lengths$toadd, labels = chrom_lengths$Chromosome) +
  theme_classic()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```
Too sensitive
TRy with segments

```{r}
CNV_segments <- CNV_segments %>% as_tibble() %>%
  mutate(chrom = factor(as.character(chrom), levels = 
                               paste0('chr', c(1:22, 'X', 'Y')))) %>%
  left_join(chrom_lengths, by = c('chrom' = 'Chromosome')) %>%
  mutate(start_genome = loc.start + toadd, 
         end_genome = loc.end + toadd,
         state = case_when(
           seg.median > 0.1 ~ 'Gain',
           seg.median < -0.1 ~ 'Loss',
           TRUE ~ 'Neutral'
         )) %>%
  arrange(ID, chrom, loc.start) %>%
  filter(!chrom %in% c('chrX', 'chrY'))

#sample index
Second <- pDat %>% filter(Trimester == 'Second') %>% pull(Sample_Name)
First <- pDat %>% filter(Trimester == 'First') %>% arrange(Case_ID) %>% pull(Sample_Name)
Third <- pDat %>% filter(Trimester == 'Third') %>% arrange(Case_ID) %>% pull(Sample_Name)


#second trim
CNV_segments %>% filter(ID %in% Second) %>%
  ggplot(aes(x =  ID, color = state)) +
  geom_segment(aes(x =  ID, xend = ID,  y = start_genome, yend = end_genome), size = 3)+
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(expand = c(0.001, 0), breaks = chrom_lengths$toadd[1:22], 
                     labels = chrom_lengths$Chromosome[1:22]) +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.minor = element_blank()) +
  coord_flip() 

#First trimester
CNV_segments %>% filter(ID %in% First[1:25]) %>%
  ggplot(aes(x =  ID, color = state)) +
  geom_segment(aes(x =  ID, xend = ID,  y = start_genome, yend = end_genome), size = 3)+
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(expand = c(0.001, 0), breaks = chrom_lengths$toadd[1:22], 
                     labels = chrom_lengths$Chromosome[1:22]) +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.minor = element_blank()) +
  coord_flip() 

CNV_segments %>% filter(ID %in% First[26:50]) %>%
  ggplot(aes(x =  ID, color = state)) +
  geom_segment(aes(x =  ID, xend = ID,  y = start_genome, yend = end_genome), size = 3)+
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(expand = c(0.001, 0), breaks = chrom_lengths$toadd[1:22], 
                     labels = chrom_lengths$Chromosome[1:22]) +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.minor = element_blank()) +
  coord_flip() 

#Third trimester
CNV_segments %>% filter(ID %in% Third[1:42]) %>%
  ggplot(aes(x =  ID, color = state)) +
  geom_segment(aes(x =  ID, xend = ID,  y = start_genome, yend = end_genome), size = 3)+
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(expand = c(0.001, 0), breaks = chrom_lengths$toadd[1:22], 
                     labels = chrom_lengths$Chromosome[1:22]) +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.minor = element_blank()) +
  coord_flip() 
CNV_segments %>% filter(ID %in% Third[43:85]) %>%
  ggplot(aes(x =  ID, color = state)) +
  geom_segment(aes(x =  ID, xend = ID,  y = start_genome, yend = end_genome), size = 3)+
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(expand = c(0.001, 0), breaks = chrom_lengths$toadd[1:22], 
                     labels = chrom_lengths$Chromosome[1:22]) +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.minor = element_blank()) +
  coord_flip() 

CNV_segments %>% filter(ID %in% Third[86:126]) %>%
  ggplot(aes(x =  ID, color = state)) +
  geom_segment(aes(x =  ID, xend = ID,  y = start_genome, yend = end_genome), size = 3)+
  scale_color_manual(values = c('Gain' = 'Red', 'Loss' = 'Blue', 'Neutral' = NA)) +
  scale_y_continuous(expand = c(0.001, 0), breaks = chrom_lengths$toadd[1:22], 
                     labels = chrom_lengths$Chromosome[1:22]) +
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        panel.grid.minor = element_blank()) +
  coord_flip() 
```

# extra stuff

```{r}
ggplot(pDat, aes(x = normalized_X_intensity, y = normalized_Y_intensity, 
                 fill = Prob_SNP_outlier)) +
  geom_point(aes(shape = Sex), size = 2.5, alpha = 0.95) +
  scale_shape_manual(values = c('M' = 21, 'F' = 22)) +
  theme_bw() +
  scale_fill_viridis_c(option = 'inferno')

```
