---
title: "15_Sample_Characteristics"
author: "Victor Yuan"
date: "29/05/2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

I realized some variables were had an incorrect class (e.g. factor when should be numeric), and some
observations needed to be replaced with NAs (0 for DNA concentration means we did not measure).

I made edits in my 11_QC script to address these issues. 
I also forgot to test for confounders / relationships between sample characteristics.


# Setup

```{r message = F, warning = F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggrepel)
library(viridis)
library(yahew)
library(egg)

pDat <- readRDS('../../data/main/interim/13_pDat.rds')
# PCA association data for plotting
pca <- readRDS('../../data/main/interim/11_pca_association_plot.rds')
```

# Relationships between variables

```{r warning = F}
glimpse(pDat)

cov_tests <- pDat %>%
  select(Row_numeric, Row_factor, Chip_number, Week,Batch_BSC,
         DNA_conc_before_load, DNA_loaded,  
         failed_probes, detP_01_minfi , beadcount, Average_intensity, 
         Case_ID, Sex, Tissue,  Trimester,
         cor_to_reference, cor_to_unrelated, Prob_SNP_outlier) %>%
  as.data.frame() %>% pairtest

# make categories
cov_tests <- cov_tests %>% 
  mutate(pval_cat = if_else(p.value < 0.001, '< 0.001',
                            if_else(p.value < 0.01, '< 0.01',
                                    if_else(p.value < 0.05, '< 0.05', '> 0.05'))))

# plot heatmap of associations
ggplot(cov_tests, aes(x=Column, y = Row, fill = pval_cat)) +
  geom_tile(col = 'grey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = c('> 0.05' = 'White', '< 0.05' = '#fee8c8', 
                               '< 0.01' = '#fdbb84', '< 0.001' = '#e34a33')) +
  labs(x = '', y = '', fill = 'p') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust= 1),
        panel.grid.major.x = element_blank()) +
  coord_equal()
```

```{r warning = F}
qplot(data = pDat,x = DNA_loaded, y = failed_probes) + geom_smooth()
qplot(data = pDat,x = DNA_conc_before_load, y = failed_probes) + geom_smooth()
qplot(data = pDat, x = Row_numeric, y = failed_probes) + geom_smooth()
qplot(data = pDat, x = Row_numeric, y = detP_01_minfi) + geom_smooth()
qplot(data = pDat, x = Week, y = detP_01_minfi, geom = 'boxplot') 
qplot(data = pDat, x = Row_numeric, y = Average_intensity) + geom_smooth()
qplot(data = pDat, x = as.numeric(as.factor(Batch_BSC)), y =Average_intensity) + 
  geom_smooth(method = 'lm') + 
  scale_x_continuous(labels = levels(as.factor(pDat$Batch_BSC))) +
  scale_y_continuous(limits = c(0, 14500))
  labs(x = 'BSC Batch')
qplot(data = pDat, x = Average_intensity, y = detP_01_minfi)
qplot(data = pDat, x = Average_intensity, y = beadcount)
qplot(data = pDat, x = Sex, y = detP_01_minfi)
ggplot(pDat, aes(x = Tissue, fill = Batch_BSC)) + geom_bar(position = 'dodge')
```

# PCA

```{r}
# create color palette
colpal <- c('white', '#fee8c8', '#fdbb84', '#e34a33')
names(colpal) <- levels(pca$Plot_data$pval_cat)

p1 <- ggplot(pca$Plot_data, aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile(col = 'lightgrey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0), labels = 1:20) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = colpal)  + 
  labs(y = '', fill = 'P value')

p1b <- ggplot(pca$PC_variance, aes(x = PC, y = Prop_var_raw)) +
  geom_bar(stat = 'identity') + theme_bw() + labs(y = '% variance') +
  scale_x_continuous(breaks = 1:20)

ggarrange(p1, p1b, heights = c(3,1))
```

```{r}
qplot(data = pDat, x = Row_numeric, y = PC5_raw)
qplot(data = pDat, x = Row_numeric, y = PC6_raw)
qplot(data = pDat, x = Row_numeric, y = PC7_raw)
qplot(data = pDat, x = Row_numeric, y = PC8_raw)
qplot(data = pDat, x = Row_numeric, y = PC10_raw)
```