---
title: "2_3_Measuring_absolute_contamination"
author: "Victor Yuan"
date: "17/06/2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

I'ved decided to analyze both first and third trimester samples together, after removing severely contaminated samples.

First it's important to consider the level of maternal contamination. Previously I hadn't been able to quantify the absolute levels of contamination. So first I develop and approach to do so, and then I do linear modelling.

# Setup

## Load libraries and data

```{r message = F, warning = F}
# libraries and data
library(ggplot2)
library(tidyverse)
library(ggrepel)
library(viridis)
library(scales)
library(RColorBrewer)
library(pheatmap)
library(irlba)
library(plomics) # 0.2 github/wvictor14/plomics
library(GGally)
library(cowplot)
library(umap)
library(dendextend)
library(stringr)
library(minfi)
library(ggridges)
library(ggpmisc)
library(egg)
```

## Data

Here I call 'syncytiotrophoblasts' into 'Trophoblasts enz' and add 'cs' suffixes to the cell sorted
samples.

```{r}
# pdata
pDat <- readRDS('../../data/main/interim/1_3_pDat.rds')

# filter to first trimester samples
pDat_first <- pDat %>%
  filter(Trimester == 'First', Tissue != 'Dead Cells and Lymphocytes') %>%
  
  # update tissue annotation
  mutate(Tissue = case_when(Tissue != 'Villi' ~ paste(Tissue, 'cs'),
                            Tissue == 'Villi' ~ Tissue))

# raw methylation data
betas <- readRDS('../../data/main/interim/1_4_betas_noob_filt.rds')
mset_noob <- readRDS('../../data/main/interim/1_4_mset_noob.rds')
colnames(mset_noob) <- colnames(betas) <- pDat$Sample_Name
mvals <- getM(mset_noob)

# filter to first trimester
mvals_first <- mvals[rownames(betas),pDat_first$Sample_Name]
betas_first <- betas[,pDat_first$Sample_Name]

# annotation
anno <- getAnnotation(mset_noob)
anno <- anno %>%
  as_tibble() %>%
  filter(Name %in% rownames(betas_first)) # filter to filtered betas cpgs
probe_anno <- readRDS('../../data/main/interim/1_1_probe_anno.rds')

# snp data
snp_betas <- readRDS('../../data/main/interim/1_1_snp_betas.rds')

# color key
color_code <- readRDS('../../data/main/interim/1_1_color_code.rds')
color_code[[1]] <- color_code[[1]] %>% 
  mutate(label = ifelse(Tissue == 'Syncytiotrophoblast', 'Trophoblasts enz', Tissue)) %>%
  mutate(label = ifelse(!label %in% c('Trophoblasts enz', 'Villi', 'Villi maternal'), 
                        paste(label, 'cs'),
                        label),
         Colors_Tissue = ifelse(label == 'Villi', '#721111', Colors_Tissue),
         Colors_Tissue = ifelse(label == 'Trophoblasts enz', '#f4702e', Colors_Tissue))
saveRDS(color_code[[1]], '../../data/main/interim/2_3_color_code.rds')

color_code_tissue <- setNames(color_code[[1]]$Colors_Tissue, color_code[[1]]$label)

```

# Contamination

## Training

These are the levels of contamination that we are dealing with:
```{r}
pDat_first %>% arrange(Tissue, cor_to_reference) %>%
  mutate(Sample_Name = factor(Sample_Name, levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = cor_to_reference)) +
  geom_point(shape = 21, aes(fill = Tissue), color = 'grey', size = 3) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_first$Tissue)]) +
  facet_wrap(~Tissue, scales = 'free_y', ncol = 1) + theme_bw() + labs(x = '') + coord_flip()
```

But we can't equate this to % contamination because we don't have a clear reference point.

We can kind of tell based using the normalized Y intensity in males, because it will go to 0 as
they get closer to 100% maternal contamination



```{r}
p_c1 <- ggplot(pDat, aes(x = cor_to_reference, y = normalized_Y_intensity, color = Sex)) +
  geom_point() + theme_bw() + stat_smooth(method = 'lm');p_c1  

ggplot(pDat, aes(x = cor_to_reference, y = normalized_Y_intensity, Shape = Sex, color = Tissue,
                 shape = Trimester)) +
  geom_point() + theme_bw()  + facet_wrap(~Tissue)
```

```{r}
fit_ycor <-lm(normalized_Y_intensity ~ cor_to_reference, pDat %>% filter(Sex == 'M'))
summary(fit_ycor)
```


`normalized_Y_intensity` is our measure of contamination for males.  With the maximum values 
representing samples that are nearly 100% fetal and 0 % contaminated with maternal DNA. The minimum 
value for males represent the most contaminated male samples. However we since we don't know the 
true levels of contamination of these samples, we can't comment if the minimum values in males 
represent 100% contamination / 0% fetal. However, instead we can say that the normalized_Y_intensity
mean for female samples can represent *what a male's normalized Y intensity value would look like if it was 100% contaminated with maternal cells*. 

Therefore, it makes sense to transform / normalize this range to:

1. Maximum -> the average `normalized_Y_intensity` for males when contamination is lowest 
2. Minimum -> the average `normalized_Y_intensity` for females.


For 1. It makes sense to use male samples with a low probability that SNPs are outliers instead of
the correlation to the reference villi. Because `cor_to_reference` assumes villi samples are 100%
not contaminated, whereas the probability of a SNP being an outlier is an unsupervised measure of
contamination. The only thing we need to be careful about is when a male sample is 100% contaminated
with maternal DNA, it's probability will also be 0, but these should be easy to spot if we look at
normalized_Y_intensity as well.

Here I look the top 20 male samples with the lowest probability of their SNPs being outliers, and 
calculate their average normalized Y intensity, this will represent our maximum value of the 
normalized range:

```{r}
# calculate minmum maximum of normalized range
pDat %>% 
  filter(Sex == 'M') %>% 
  arrange(Prob_SNP_outlier) %>% 
  select(Sample_Name, Sex, Prob_SNP_outlier, normalized_Y_intensity) %>% 
  
  # display top 30
  dplyr::slice(1:20) %>%
  as.data.frame() %>% {
    ggplot(., aes(x = Prob_SNP_outlier, y = normalized_Y_intensity)) +
      geom_smooth(method = 'lm') + geom_point() +
      geom_hline(., yintercept = mean(.$normalized_Y_intensity), 
                 linetype = 'dashed', color = 'red') +
      geom_text(x = min(.$Prob_SNP_outlier) + 0.0005, y = mean(.$normalized_Y_intensity) + 0.01,
                label = paste0('mean = ', formatC(mean(.$normalized_Y_intensity)), digits = 2), 
                col = 'red') +
      theme_bw()
}

max_Y <- pDat %>% 
  filter(Sex == 'M') %>% 
  arrange(Prob_SNP_outlier) %>%
  dplyr::slice(1:20) %>%
  pull(normalized_Y_intensity) %>%
  mean()
```

Now for 1., the minimum of our normalized range will be the average female y intensity:

```{r}
min_Y <- pDat %>% filter(Sex == 'F') %>%
  pull(normalized_Y_intensity) %>%
  mean()

min_Y
```

## Predict on females

To calculate our measure of contamination for females:

1. predict normalized_Y_intensity based on cor_to_reference using model trained in males
2. Normalize to 0 and 1, based on min and max
3. Samples falling outside of min and max are set to 0 and 1 respectively
4. Reverse the scale such that 1 is 100% maternal contamination and 0 is 0%

For females, we skip step 1, and do step 2. and 3.


```{r}
pDat <- pDat %>% 
  
  # 1.
  ## predict values for females
  mutate(maternal_contamination = predict(fit_ycor, pDat %>% select(cor_to_reference))) %>%
  
  ## set as normalized_y_intensity for males
  mutate(maternal_contamination = ifelse(Sex == 'M', normalized_Y_intensity, maternal_contamination)) %>%
  

  mutate(maternal_contamination_norm = case_when(
    
    #2.
    maternal_contamination > max_Y ~ 1,
    maternal_contamination < min_Y ~ 0,
    
    #3.
    TRUE ~ ((maternal_contamination - min_Y) / (max_Y - min_Y))
  )) %>%
  
  #4.
  mutate(maternal_contamination_norm_flip = 1-maternal_contamination_norm)

pDat %>%
  select(Sex, contains('maternal_contamination'), cor_to_reference, normalized_Y_intensity) %>%
  gather(key = step, value = Contamination, -Sex, -cor_to_reference) %>%
  mutate(step = factor(case_when(
    step == 'normalized_Y_intensity' ~ '1. Raw data',
    step == 'maternal_contamination' ~ '2. Predicted value for females\nTrained values for males',
    step == 'maternal_contamination_norm' ~ '3. Normalized and constrained to 0 and 1',
    step == 'maternal_contamination_norm_flip' ~ '4. Reversed, such that 1 = 100%'
  ), levels = c('1. Raw data', 
                '2. Predicted value for females\nTrained values for males',
                '3. Normalized and constrained to 0 and 1', 
                '4. Reversed, such that 1 = 100%'))) %>% 
  {
    ggplot(., aes(x = cor_to_reference, y = Contamination, color = Sex)) +
      geom_smooth(data = . %>% filter(step == '1. Raw data', Sex == 'M'), method = 'lm',
                  show.legend = F) +
      geom_abline(data = data.frame(slope = fit_ycor$coefficients[2], 
                                    intercept = fit_ycor$coefficients[1],
                                    step = c('2. Predicted value for females\nTrained values for males'),
                                    Sex = c('M')),
                  aes(slope = slope, intercept = intercept, color = Sex)) +
      geom_point(alpha = 0.7, size = 1.5) +
      facet_wrap(~step, ncol = 2) +
      theme_bw()
}

```

One thing to note is that there are a large number of samples that are estimated to have 
contamination > 50%

```{r}
pDat %>% arrange(Tissue, maternal_contamination_norm_flip) %>% 
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  gather(key = Contamination, value = value, maternal_contamination_norm_flip, Prob_SNP_outlier) %>%
  ggplot(aes(x = Sample_Name, y = value, color = Tissue)) +
  geom_point() + 
  geom_segment(aes(y = 0, yend = value, 
                   x = Sample_Name, xend = Sample_Name))+
  scale_color_brewer(palette = 'Set1') +
  scale_y_continuous(limits = c(0,1), expand = c(0,0), breaks = seq(0,1, 0.1)) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  facet_grid(Contamination~Trimester, scales = 'free_x', space = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) 

ggplot(pDat, aes(y = maternal_contamination_norm_flip, x = Prob_SNP_outlier)) +
  geom_point() 

pDat %>% arrange(Tissue, maternal_contamination_norm_flip) %>% 
  filter(maternal_contamination_norm_flip < 0.5) %>%
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = maternal_contamination_norm_flip, color = Tissue)) +
  geom_point() + 
  geom_segment(aes(y = 0, yend = maternal_contamination_norm_flip, 
                   x = Sample_Name, xend = Sample_Name))+
  scale_color_brewer(palette = 'Set1') +
  scale_y_continuous(limits = c(0,1), expand = c(0,0), breaks = seq(0,1, 0.1)) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  facet_grid(.~Trimester, scales = 'free_x', space = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) 

ggplot(pDat, aes(y = maternal_contamination_norm_flip, x = Prob_SNP_outlier)) +
  geom_point() 


# join to subsets first rtimester pdat
pDat_first <- pDat_first %>%
  left_join(pDat %>% select(Sample_Name, maternal_contamination_norm_flip))
```

maternal_contamination_norm_flip: final measure of absolute contamination

## Predict on mat blood

First I define the function of our trained model

```{r}
# trained parameters
intercept <- fit_ycor$coefficients[1]
slope <- fit_ycor$coefficients[2]

infer_maternal <- function(x) {
  
  # 1. predict normalized_Y_intensity based on cor_to_reference using model trained in males
  y <- slope*x + intercept
  
  # 2 & 3 Normalize to 0 and 1, based on min and max, constrain to min max
  y_norm <- 
    
    case_when(
      #2.
      y > max_Y ~ 1,
      y < min_Y ~ 0,
      
      #3.
      TRUE ~ ((y - min_Y) / (max_Y - min_Y))
    )
  
  # 4. Reverse the scale such that 1 is 100% maternal contamination and 0 is 0%
  y_norm_flip <- 1 - y_norm
  
  return(y_norm_flip)
}

# check that this matches our previous manual calculation
pDat %>% filter(Sex == 'F') %>%
  mutate(out = infer_maternal(cor_to_reference)) %>%
  select(out, maternal_contamination_norm_flip)
```

```{r}
ss <- readxl::read_xlsx('Z:/ROBLAB6 Infinium450k John/Master_Sample_Sheet.xlsx')

table(ss$Tissue)

ss_matched <- ss %>% 
  filter(Platform != '27k') %>%
  group_by(Case_ID) %>%
  filter(any(Tissue %in% c('Maternal Blood', 'Amnion', 'Chorion', 'Decidua')) & 
           any(Tissue == 'Villi')) %>%
  ungroup() %>% arrange(desc(Platform), Case_ID, Tissue) %>%
  
  # remove duplicated rows
  group_by(Basename) %>% 
  dplyr::slice(1) %>%
  ungroup()

rgset_matched_450k <- read.metharray.exp(targets = ss_matched %>% filter(Platform == '450k'))
rgset_matched_850k <- read.metharray.exp(targets = ss_matched %>% filter(Platform == '850k'))

snps_matched_450k <- getSnpBeta(rgset_matched_450k)
snps_matched_850k <- getSnpBeta(rgset_matched_850k)

colnames(snps_matched_450k) <- ss_matched %>% filter(Platform == '450k') %>% pull(Sample_Name)
colnames(snps_matched_850k) <- ss_matched %>% filter(Platform == '850k') %>% pull(Sample_Name)

# calculate for all pairwise comparisons
cor(snps_matched_450k)

res_450k <- cor(snps_matched_450k) %>% as_tibble() %>% 
  mutate(Sample1 = colnames(.)) %>% 
  
  # add case_ID and tissue
  left_join(ss_matched %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  
  # longate the dataframe, each row is a single pairwise comparison
  gather(key = 'Sample2', value = 'Cor', -Sample1, -Case_ID, -Tissue) %>%
  
  # pData needs to refer to the right sample (1 or 2)
  dplyr::rename(Case_ID1 = Case_ID, Tissue1 = Tissue) %>%
  
  # add info for sample 2
  left_join(ss_matched %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample2' = 'Sample_Name')) %>%
  dplyr::rename(Case_ID2 = Case_ID, Tissue2 = Tissue) %>% 
  
  # only interested in within-case comparisons, not to each other
  filter(Case_ID1 == Case_ID2, Sample1 != Sample2,
         Tissue1 == 'Villi', Tissue2 != 'Villi') %>%
  
  # for samples with more than one villi, take the mean correlation to all of theme
  group_by(Sample2) %>%
  summarize(cor_to_villi = mean(Cor)) %>%
  
  left_join(ss_matched %>% select(Sample_Name, Tissue, Case_ID), by = c('Sample2' = 'Sample_Name')) %>%
  
  # calculat maternal contam
  mutate(maternal_contamination = infer_maternal(cor_to_villi))

# repeat for 850k
res_850k <- cor(snps_matched_850k) %>% as_tibble() %>% 
  mutate(Sample1 = colnames(.)) %>% 
  left_join(ss_matched %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample1' = 'Sample_Name')) %>%
  gather(key = 'Sample2', value = 'Cor', -Sample1, -Case_ID, -Tissue) %>%
  dplyr::rename(Case_ID1 = Case_ID, Tissue1 = Tissue) %>%
  left_join(ss_matched %>% select(Sample_Name, Case_ID, Tissue), by = c('Sample2' = 'Sample_Name')) %>%
  dplyr::rename(Case_ID2 = Case_ID, Tissue2 = Tissue) %>% 
  filter(Case_ID1 == Case_ID2, Sample1 != Sample2,
         Tissue1 == 'Villi', Tissue2 != 'Villi') %>%
  group_by(Sample2) %>%
  summarize(cor_to_villi = mean(Cor)) %>%
  left_join(ss_matched %>% select(Sample_Name, Tissue, Case_ID), by = c('Sample2' = 'Sample_Name')) %>%
  mutate(maternal_contamination = infer_maternal(cor_to_villi))

res <- bind_rows(res_850k, res_450k) 

res %>%
  ggplot(aes(x = Sample2, y = maternal_contamination, color = Tissue)) +
  geom_point(size = 3) +
  geom_segment(aes(y = 0, yend = maternal_contamination, 
                   x = Sample2, xend = Sample2), size = 1) +
  geom_text(data = . %>%
              mutate(label = formatC(maternal_contamination, 2, format = 'f')),
            aes(label = label), nudge_y = 0.05, show.legend = F) +
  theme_bw() + 
  coord_flip()

```

TODO: KEEP THIS SEPARATE FROM DMC ANALYSIS


