---
title: "2_4_combined_analysis"
author: "Victor Yuan"
date: "08/07/2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

# Setup

## Libraries

```{r}
# libraries and data
library(ggplot2)
library(tidyverse)
library(ggrepel)
library(viridis)
library(scales)
library(RColorBrewer)
library(pheatmap)
library(irlba)
library(plomics) # 0.2 github/wvictor14/plomics
library(GGally)
library(cowplot)
library(umap)
library(dendextend)
library(stringr)
library(minfi)
library(ggridges)
library(ggpmisc)
library(egg)
library(limma)
library(broom)
library(missMethyl)
library(kableExtra)
```

## Data

```{r}
pDat <- readRDS('../../data/main/interim/2_3_pDat_contam.rds')
pDat <- pDat %>%
  mutate(Tissue = case_when(
    !(Tissue %in% c('Villi', 'Villi maternal', 'Syncytiotrophoblast')) ~ paste(Tissue, 'cs'),
    Tissue == 'Syncytiotrophoblast' ~ 'Trophoblasts enz',
    TRUE ~ Tissue
  )) 

# raw methylation data
betas <- readRDS('../../data/main/interim/1_4_betas_noob_filt.rds')

mset_noob <- readRDS('../../data/main/interim/1_4_mset_noob.rds') # for mvals
colnames(mset_noob) <- colnames(betas) <- pDat$Sample_Name
mvals <- getM(mset_noob)

# annotation
anno <- getAnnotation(mset_noob)
anno <- anno %>%
  as_tibble() %>%
  filter(Name %in% rownames(betas_first)) # filter to filtered betas cpgs
probe_anno <- readRDS('../../data/main/interim/1_1_probe_anno.rds')

# color key
pheatmap_color_code <- readRDS('../../data/main/interim/1_1_color_code.rds')

color_code <- readRDS('../../data/main/interim/2_3_color_code.rds')
color_code_tissue <- setNames(color_code$Colors_Tissue, color_code$label)
```

## Remove contaminated samples

Her eI remove samples with > 35% maternal contamination based on previous analyses.


```{r}
#Before contamination
pDat %>%
  filter(!Tissue %in% c('Villi maternal', 'Trophoblasts enz', 'Mixture cs', 'Dead Cells and Lymphocytes cs')) %>%
  arrange(Tissue, maternal_contamination_norm_flip) %>% 
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  ggplot(aes(x = Sample_Name, y = maternal_contamination_norm_flip, color = Tissue)) +
  geom_point() + 
  geom_segment(aes(y = 0, yend = maternal_contamination_norm_flip, 
                   x = Sample_Name, xend = Sample_Name))+
  scale_color_manual(values = color_code_tissue) +
  scale_y_continuous(limits = c(0,1), expand = c(0,0), breaks = seq(0,1, 0.1)) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  facet_grid(Trimester~., scales = 'free_y', space = 'free_y') +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank())  +
  labs(x = 'Samples', y = 'Maternal contamination') +
  coord_flip() +
  geom_hline(yintercept = 0.35, col = 'red', linetype = 'dashed')

# after
pDat %>%
  filter(!Tissue %in% c('Villi maternal', 'Trophoblasts enz', 'Mixture cs', 'Dead Cells and Lymphocytes cs')) %>%
  arrange(Tissue, maternal_contamination_norm_flip) %>% 
  mutate(Sample_Name = factor(as.character(Sample_Name), levels = Sample_Name)) %>%
  filter(maternal_contamination_norm_flip < 0.35) %>%
  ggplot(aes(x = Sample_Name, y = maternal_contamination_norm_flip, color = Tissue)) +
  geom_point() + 
  geom_segment(aes(y = 0, yend = maternal_contamination_norm_flip, 
                   x = Sample_Name, xend = Sample_Name))+
  scale_color_manual(values = color_code_tissue) +
  scale_y_continuous(limits = c(0,1), expand = c(0,0), breaks = seq(0,1, 0.1)) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  facet_grid(Trimester~., scales = 'free_y', space = 'free_y') +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank())  +
  labs(x = 'Samples', y = 'Maternal contamination') +
  coord_flip() +
  geom_hline(yintercept = 0.35, col = 'red', linetype = 'dashed')

#summarize filtering
x <- pDat %>%
  group_by(Trimester, Tissue) %>%
  summarize(Before_filter = n())
y <- pDat %>%
  filter(maternal_contamination_norm_flip < 0.35) %>%
  group_by(Trimester, Tissue) %>%
  summarize(After_filter = n())
removed_df <- x %>% 
  ungroup %>%
  mutate(After_filter = y$After_filter,
         Number_removed = Before_filter-After_filter)

removed_df %>%
  kable(align = 'c', escape = F) %>%
  kable_styling(c("striped", "condensed"),full_width = F) %>%
  row_spec(which(removed_df$Number_removed>0), bold = T) 
```

## Remove outliers

Previsouly had flagged term and 1st trimester samples with odd methylation profiles. Here I remove
these.

I previously stated in 2_1_Term_contamination.Rmd that I would remove hofbauer samples PM366, PM381 due
to contamination. However these are all below our contamination threshold. Therefore I just remove
PM364, since this ample appeared to be highly 'trophoblastic' with respect to DNA methylation.


I also remove: 
* tissues that I won't look at for this analysis.
* replicates
*P131_hofb_cs as we think is contaminated with endo/stromal cells


```{r}
pDat_filt <- pDat %>% 
  filter(maternal_contamination_norm_flip < 0.35,
         !Sample_Name %in% c('PM364_hofb_cs', 'PL293_v_R2', 'PM366_vc_R2', 'P131_hofb_cs'),
         !Tissue %in% c('Villi maternal', 'Trophoblasts enz', 'Mixture cs', 'Dead Cells and Lymphocytes cs'))

# filter to first trimester
mvals_filt <- mvals[rownames(betas),pDat_filt$Sample_Name]
betas_filt <- betas[,pDat_filt$Sample_Name]
```


# 1.0 Global methylation

## Density distributions

Here we plot the density distributions of the betas across each tissue

Density for each sample:

```{r}
#calculate densities
sample_densities <- pDat_filt %>%
  select(Sample_Name, Tissue) %>%
  mutate(densities = apply(betas_filt, 2, density)) %>%
  mutate(x = map(densities, 'x'),
         y = map(densities, 'y')) %>%
  select(-densities) %>%
  unnest() %>%
  # add Trimester information
  left_join(pDat_filt %>% select(Sample_Name, maternal_contamination_norm_flip, Trimester)) 

# plot all densities
sample_densities %>%
  filter(Trimester != 'Second') %>%
  
  # plot
  ggplot(aes(x = x, y = y, color = Tissue, group = Sample_Name)) +
  geom_line(size = 1, alpha = 0.5) + theme_bw() +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 25, 50, 75, 100))  +
  facet_grid(Trimester~Tissue) +
  scale_color_manual(values = color_code_tissue[unique(pDat_filt$Tissue)]) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  labs(x = '% methylation', y = 'density')

# just trophoblast from first
sample_densities %>%
  filter(Tissue == 'Trophoblasts cs', Trimester == 'First') %>%
  ggplot(aes(x = x, y = y, color = Sample_Name, group = Sample_Name)) +
  geom_line(size = 1, alpha = 0.5) + theme_bw()  +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) 

sample_densities %>%
  filter(Tissue == 'Trophoblasts cs', Trimester == 'First') %>%
  ggplot(aes(x = x, y = y, color = maternal_contamination_norm_flip, group = Sample_Name)) +
  geom_line(size = 1, alpha = 0.5) + theme_bw()  +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) 
```

PL290 has an odd peak in the middle

## PCA

```{r}
# compute pca
set.seed(1)
pca <- prcomp_irlba(t(betas_filt), n = 20, center = T, scale = F)

# add pc scores to pdata
pca_scores <- pca$x[,1:20] %>% as.data.frame()
colnames(pca_scores) <- paste0(colnames(pca_scores), '_processed')
pDat_filt <- pDat_filt %>% 
  bind_cols(pca_scores)

# create proportion variance explained data frame
pc_info <- summary(pca)$importance %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = 'variable') %>%
  gather(key = 'PC', value = 'value', -variable) %>%
  as_tibble() %>%
  mutate(PC = factor(as.character(PC), levels = paste0('PC', 1:20)),
         Label = ifelse(variable == 'Proportion of Variance',
                        paste0(PC, ' (', prettyNum(value*100, digits = 2), '%)'),
                        as.character(PC))) %>%
  arrange(variable, PC)

  
# correlate PCs with phenodata
pc_cor <- lmmatrix(dep = pca$x[,1:20, drop = F],
                   ind = pDat_filt %>%
                   dplyr::select(Case_ID, Tissue, Sex, Trimester, #bio
                                 Week, Chip_number, Row_numeric, Row_factor, Batch_BSC, # batch
                                 DNA_loaded, 
                                 failed_probes,
                                 maternal_contamination_norm_flip),
                   metric = 'Pvalue')
# plot data
pc_cor <- pc_cor %>% as.data.frame() %>% 
  
  # add dep variables
  mutate(dep = rownames(pc_cor)) %>%
  
  # reshape
  gather(PC, pval, -dep) %>%
  
  # pvalue categories
  mutate(pval_cat = factor(case_when(
    pval > 0.05  ~ '> 0.05',
    pval < 0.05 & pval > 0.01 ~ '< 0.05',
    pval < 0.01 & pval > 0.001 ~ '< 0.01',
    pval < 0.001 ~ '< 0.001'
  ), levels = c('> 0.05', '< 0.05','< 0.01', '< 0.001')),
  
  # make PC is encoded with proper levels!!!
  PC = factor(PC, levels = paste0('PC', 1:20))) %>% as_tibble()

# create color palette
colpal <- c('white', '#fee8c8', '#fdbb84', '#e34a33')
names(colpal) <- levels(pc_cor$pval_cat)

# relevel
pc_cor <- pc_cor %>% 
  mutate(dep = factor(dep, levels = c('Case_ID', 'Tissue', 'Sex', 'Trimester',
                                     'Week', 'Chip_number', 'Row_numeric', 'Row_factor', 
                                     'Batch_BSC', # batch
                                     'DNA_loaded',
                                     'failed_probes',
                                     'maternal_contamination_norm_flip')))
                                     
p7 <- ggplot(pc_cor, aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile(col = 'lightgrey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0), labels = 1:20) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = colpal)  + 
  labs(y = '', fill = 'P value')

p8 <- ggplot(pc_info %>% filter(variable == 'Proportion of Variance') %>%
               mutate(value = value*100), 
             aes(x = PC, y = value)) +
  geom_bar(stat = 'identity') +
  theme_bw() + 
  scale_x_discrete(expand = c(0, 0), labels = 1:20) + 
  scale_y_continuous(expand = c(0, 0)) +
  labs(y = '% variance explained')

egg::ggarrange(p7, p8, heights = c(3,1))


# create another without all variables
var_interest <- c('Tissue', 'Sex', 'Trimester', 'maternal_contamination_norm_flip','Case_ID',
                  'Week', 'Chip_number', 'Row_numeric', 'Row_factor', 'Batch_BSC')

p9 <- pc_cor %>% 
  filter(dep %in% var_interest) %>%
  mutate(dep = ifelse(dep %in% var_interest[6:11], paste0('TECH_', dep), as.character(dep)),
         dep = gsub('maternal_contamination_norm_flip', 'Percent_maternal', dep)) %>%
  mutate(dep = factor(dep, levels = c('TECH_Week', 'TECH_Chip_number', 'TECH_Row_numeric',
                                      'TECH_Row_factor', 'TECH_Batch_BSC', 
                                      'Percent_maternal',  'Sex',
                                      'Case_ID', 'Trimester', 'Tissue'))) %>%
  ggplot(aes(x = PC, y = dep, fill = pval_cat)) +
  geom_tile(col = 'lightgrey') + theme_bw() +
  scale_x_discrete(expand = c(0, 0), labels = 1:20) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_fill_manual(values = colpal)  + 
  labs(y = '', fill = 'P value')

egg::ggarrange(p9, p8, heights = c(3,1))
```

Scatterplots

```{r message = F}
scatter <- function(x, y, fill, point_size = 1){
  xlab <- pc_info %>% filter(variable == 'Proportion of Variance', PC == x) %>% pull(Label)
  ylab <- pc_info %>% filter(variable == 'Proportion of Variance', PC == y) %>% pull(Label)
  
  x <- paste0(x, '_processed')
  y <- paste0(y, '_processed')
  
  out <- ggplot(pDat_filt, aes_string(x = x, y = y, fill = fill, shape = 'Trimester')) +
    geom_point(size = point_size) + theme_bw() + labs(x = xlab, y = ylab) +
    scale_shape_manual(values = c('First' = 24, 'Second' = 22, 'Third' = 21))+
    guides(fill=guide_legend(override.aes=list(shape=21)))
  
  if (is.numeric(as.data.frame(pDat_filt)[,fill])){
    out <- out +
      scale_fill_viridis()
  } else {
    out <- out + 
      scale_fill_brewer(palette = 'Set1')
  }
    
  out
}

scatter(x = 'PC1', y = 'PC2', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_filt$Tissue)])  +
  labs(fill = '') 

scatter(x = 'PC1', y = 'PC2', fill = 'maternal_contamination_norm_flip', point_size = 2) +
  scale_fill_viridis()  +
  labs(fill = '') 

scatter(x = 'PC3', y = 'PC4', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_filt$Tissue)]) +
  labs(fill = '') +
  geom_text_repel(data = pDat_filt %>% filter(Tissue == 'Trophoblasts cs', 
                                              PC4_processed < 50,
                                              Trimester == 'First'),
                  aes(x = PC3_processed, y = PC4_processed, label = Sample_Name)) 

scatter(x = 'PC3', y = 'PC4', fill = 'maternal_contamination_norm_flip', point_size = 2) +
  scale_fill_viridis()  +
  labs(fill = '') 

scatter(x = 'PC5', y = 'PC6', fill = 'Tissue', point_size = 2) +
  scale_fill_manual(values = color_code_tissue[unique(pDat_filt$Tissue)]) +
  labs(fill = '')

```

## Pairwise Correlation

```{r}
cor_betas <- cor(betas_filt)

# annotation bar
annotation <- pDat_filt %>%
  select(Tissue, Trimester, Sex, maternal_contamination_norm_flip) %>% as.data.frame()
rownames(annotation) <- pDat_filt$Sample_Name

#colors for heatmap
anno_colors <- list(
  Tissue = color_code_tissue[unique(pDat_filt$Tissue)],
  Sex = setNames(pheatmap_color_code[[2]]$Colors_Sex, pheatmap_color_code[[2]]$Sex),
  Trimester = setNames(pheatmap_color_code[[3]]$Colors_Trimester, pheatmap_color_code[[3]]$Trimester),
  maternal_contamination_norm_flip = viridis(n = 3, option = 'B')
)


# pheatmap
pheatmap(cor_betas, annotation_col = annotation, annotation_row = annotation[,'Tissue', drop = F],
         show_rownames = F, show_colnames = T,
         annotation_colors = anno_colors,
         labels_col = '',
         color = viridis(100, option = 'B'),
         cutree_cols = 5)
```

Effect of correlation to reference and stromal cs is not associated with any of the top 6 pcs, so
I will not adjust for it

# 2.0 Linear modelling

## With  vs w/o adjustment

..for maternal contamination,


```{r}
# create a design matrix
design <- pDat_filt  %>% select(Tissue, Case_ID, cor_to_reference) %>%
  mutate(Tissue = gsub('Endothelial cs', 'Endo_cs',
                       gsub('Hofbauer cs', 'Hofb_cs',
                            gsub('Trophoblasts cs', 'Troph_cs',
                                 gsub('Stromal cs', 'Strom_cs',
                                      gsub(':', '\\.', 
                                           Tissue))))))

# specify the interactions between contam and specific tissues
design <- model.matrix(~0 + Tissue+ cor_to_reference + Case_ID , 
                       data = design)

colnames(design) <- gsub(':', '.', gsub('Tissue', '', gsub('Case_ID', '', colnames(design)))) # rename columns
design <- design[,setdiff(colnames(design),'Strom_cs.cor_to_reference')]

# fit the linear model 
fit_m <- lmFit(mvals_first_filt, design)

# fit contrasts
contMatrix_1vAll <- makeContrasts(Endo_cs - (Hofb_cs + Strom_cs + Troph_cs)/3,
                            Hofb_cs - (Endo_cs + Strom_cs + Troph_cs)/3,
                            Strom_cs - (Endo_cs + Hofb_cs + Troph_cs)/3,
                            Troph_cs - (Endo_cs + Hofb_cs + Strom_cs)/3,
                            levels=design)

fit_m <- contrasts.fit(fit_m, contMatrix_1vAll)
fit_m <- eBayes(fit_m) # moderate t stats

# pull out dmcs
dmcs <- broom::tidy(fit_m) %>%
  separate(term, into = c('Group1', 'Group2'), sep = '\\s-\\s', remove = F) %>%
  group_by(Group1, Group2) %>%
  mutate(fdr = p.adjust(p.value, method = "fdr"),
         bonferroni = p.adjust(p.value, method = 'bonferroni')) %>%
  ungroup() %>%
  select(-term, -Group2)

# add delta betas
fit_b <- lmFit(betas_first_filt, design) %>% 
  contrasts.fit(contMatrix_1vAll) %>%
  eBayes() %>%
  tidy() %>%
  separate(term, into = c('Group1', 'Group2'), sep = '\\s-\\s', remove = F) %>%
  dplyr::rename(delta_b = estimate) %>%
  select(gene, Group1, delta_b)
  
dmcs <- dmcs %>% left_join(fit_b) 
```

Without adjustment:

```{r}
# repeat without adjusting for contamination
design <- pDat_first_filt  %>% select(Tissue, Case_ID, cor_to_reference) %>%
  mutate(Tissue = gsub('Endothelial cs', 'Endo_cs',
                       gsub('Hofbauer cs', 'Hofb_cs',
                            gsub('Trophoblasts cs', 'Troph_cs',
                                 gsub('Stromal cs', 'Strom_cs',
                                      gsub(':', '\\.', 
                                           Tissue))))))

design_noadj <- model.matrix(~0+Tissue + Case_ID, data = design)
colnames(design_noadj) <- gsub(':', '\\.', 
                               gsub('Tissue', '',
                                    gsub('Case_ID', '', colnames(design_noadj)))) # rename columns

contMatrix_1vAll_noadj <- makeContrasts(Endo_cs - (Hofb_cs + Strom_cs + Troph_cs)/3,
                            Hofb_cs - (Endo_cs + Strom_cs + Troph_cs)/3,
                            Strom_cs - (Endo_cs + Hofb_cs + Troph_cs)/3,
                            Troph_cs - (Endo_cs + Hofb_cs + Strom_cs)/3,
                            levels=design_noadj)

# fit the linear model ☺
fit_m_noadj <- lmFit(mvals_first_filt, design_noadj) %>%
  contrasts.fit(contMatrix_1vAll_noadj) %>%
  eBayes() %>%
  broom::tidy() %>%
  separate(term, into = c('Group1', 'Group2'), sep = '\\s-\\s') %>%
  group_by(Group1) %>%
  mutate(fdr = p.adjust(p.value, method = "fdr"),
         bonferroni = p.adjust(p.value, method = 'bonferroni'))  %>%
  ungroup() %>%
  select(-Group2)

# add delta betas
fit_b_noadj <- lmFit(betas_first_filt, design_noadj) %>%
  contrasts.fit(contMatrix_1vAll_noadj) %>%
  eBayes() %>%
  tidy() %>%
  separate(term, into = c('Group1', 'Group2'), sep = '\\s-\\s') %>%
  dplyr::rename(delta_b = estimate) %>%
  select(gene, Group1, delta_b)

dmcs_noadj <- fit_m_noadj %>% left_join(fit_b_noadj)
```

compare: 

```{r}
effects <- full_join(
  dmcs %>% group_by(Group1) %>%
    summarize(with_adj = sum(fdr < 0.01 & abs(delta_b) > 0.1)),
  dmcs_noadj %>% group_by(Group1) %>%
    summarize(without_adj = sum(fdr < 0.01 & abs(delta_b) > 0.1))) %>%
  gather(key = key, value = count, -Group1) 

effects %>% 
  ggplot(aes(x = key, y = count, color = Group1)) +
  geom_point() +
  geom_line(aes(group = Group1)) + theme_bw() +
  scale_x_discrete(labels = c('w/ adjustment', 'w/o adjustment')) +
  scale_y_continuous(breaks = seq(0,350000, 50000), labels = seq(0,350000, 50000)/10000) +
  labs(y = 'Number CpGs in 10,000s', x = '', title = 'Effect of adjusting for contamination')
```

# Save Data

```{r, eval = F}
saveRDS(dmcs, '../../data/main/interim/2_4_dmcs_with_adjustment.rds')
saveRDS(effects, '../../data/main/interim/2_4_effects_of_adjustment.rds')
```

# SessionInfo
`r sessioninfo()`

