---
title: "2_7_PMDs"
author: "Victor Yuan"
date: "23/09/2019"
output:
  html_document:
    df_print: paged
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

# Setup

## Libraries

```{r, message = FALSE, warning = FALSE}
# libraries and data
library(tidyverse); theme_set(theme_bw())
library(kableExtra)
library(glue)
library(scales)
library(rtracklayer) 
library(DT)
library(readxl)
library(fuzzyjoin)
```

## Data

```{r, message = FALSE, warning = FALSE}
pDat <- readRDS('../../data/main/interim/2_3_pDat_contam.rds')
pDat <- pDat %>%
  mutate(Tissue = case_when(
    !(Tissue %in% c('Villi', 'Villi maternal', 'Syncytiotrophoblast')) ~ paste(Tissue, 'cs'),
    Tissue == 'Syncytiotrophoblast' ~ 'Trophoblasts enz',
    TRUE ~ Tissue
  )) 

# raw methylation data
betas <- readRDS('../../data/main/interim/1_4_betas_noob_filt.rds')

mset_noob <- readRDS('../../data/main/interim/1_4_mset_noob.rds') # for mvals
colnames(mset_noob) <- colnames(betas) <- pDat$Sample_Name
mvals <- getM(mset_noob)

probe_anno <- readRDS('../../data/main/interim/1_1_probe_anno.rds')

# color key
pheatmap_color_code <- readRDS('../../data/main/interim/1_1_color_code.rds')

color_code <- readRDS('../../data/main/interim/2_3_color_code.rds')
color_code_tissue <- setNames(color_code$Colors_Tissue, color_code$label)

# DMCs
dmcs <- readRDS('../../data/main/interim/2_4_dmcs.rds')

# annotation
anno <- readRDS('Z:/Victor/Repositories/EPIC_annotation/hg19_epic_annotation.rds')

#color code
color_code <- readRDS('../../data/main/interim/2_3_color_code.rds')
color_code_tissue <- setNames(color_code$Colors_Tissue, color_code$label)

# 450k annotation
anno_450k <- read_csv('Z:/Victor/Data/DNAm annotations/HumanMethylation450_15017482_v1-2.csv',
                      skip = 7)

# all cpgs in the genome
hg19_cpgs <- readRDS('../../data/main/processed/0_2_all_cpgs_hg19.rds')

#pmds
pmds <- read_xls('Z:/Victor/Projects/NIH - cells/data/external/placental_pmds.xls', col_names = F)
colnames(pmds) <- c('chr', 'start', 'end')

```

## Remove samples

remove contamined and non-interesting samples

```{r}
pDat_filt <- pDat %>% 
  filter(maternal_contamination_norm_flip < 0.35,
         !Sample_Name %in% c('PM364_hofb_cs', 'PL293_v_R2', 'PM366_vc_R2', 'P131_hofb_cs'),
         !Tissue %in% c('Villi maternal', 'Trophoblasts enz', 'Mixture cs', 'Dead Cells and Lymphocytes cs'))

# filter to first trimester
mvals_filt <- mvals[rownames(betas),pDat_filt$Sample_Name]
betas_filt <- betas[,pDat_filt$Sample_Name]
```

# PMD analysis

## Filter cpgs

D.I schroeder state that they remove probes in  the following regions:

* promoters
* cpg islands
* cpg island shores

So here I compare this filter to an unfiltered set of cpgs, and also 450k / EPIC

A table of number of cpgs per filtered/ unfiltered, and 450k/epic cpgs:

```{r}
# different sets of cpgs
cpg_sets <- expand_grid(filt = c('unfiltered', 'filtered'), 
                        array = c('450k', 'EPIC'))

# all cpgs in pmds (unfiltered, filtered)
cpg_pmd_unfilt <- anno %>%
  filter(!is.na(pmd_id),
         cpg %in% rownames(betas_filt)) 
cpg_pmd_filt <- cpg_pmd_unfilt %>%
  filter(!is.na(pmd_id),
         !cpg_id %in% c('island', 'shore'),
         !grepl('promoter', genes_id))

# get cpgs per set
cpg_sets <- cpg_sets %>%
  mutate(cpg = case_when(
    filt == 'unfiltered' & array == 'EPIC' ~ list(cpg_pmd_unfilt$cpg),
    filt == 'unfiltered' & array == '450k' ~ list(intersect(cpg_pmd_unfilt$cpg, anno_450k$IlmnID)),
    
    filt == 'filtered' & array == 'EPIC' ~ list(cpg_pmd_filt$cpg),
    filt == 'filtered' & array == '450k' ~ list(intersect(cpg_pmd_filt$cpg, anno_450k$IlmnID))
  ))

#display number of cpgs
cpg_sets %>%
  mutate(n = map_int(cpg, length)) %>%
  select(-cpg) %>%
  pivot_wider(
    id_cols = filt,
    names_from = array,
    values_from = n
  )
```

## Calculate densities

Now I can calculate the density per celltype / trimeseter. I follow the same code from my enrichment
analysis.

I create a new column in `cpg_sets` that is a nested data frame that corresponds to the density over the betas for each celltype/trimester across cpgs in each cpg set.

```{r}
# get betas in pmds, by cell type and trimester,
betas_pmds <- cpg_sets %>%
  
  # add a data-frame column that is the pDat with a nested sample names column
  mutate(pdat = list(
           pDat_filt %>%
  
           # get sample names for each group
           select(Trimester, Tissue, Sample_Name) %>%
           dplyr::rename(Celltype = Tissue) %>%
           group_by(Trimester, Celltype) %>%
           summarize(Samples =list(Sample_Name))
           )) %>%
  
  # get betas for each group
  unnest(pdat) %>%
  mutate(betas = 
           map2(cpg, Samples,
                ~ as.vector(betas_filt[.x, .y]))) %>%
  select(-cpg, -Samples)

 # calculate density
densities_pmds <- betas_pmds %>% 
  mutate(density = map(betas, density),
         x = map(density, 'x'),
         y = map(density, 'y')) %>%
  
  # clean up
  select(-betas, -density) %>%
  unnest(c(x, y))


# calculate percent methylation over regions
counts_pmds <- betas_pmds %>%
  unnest(betas) %>%
  mutate(range = cut(betas, 
                     c(-Inf, 0.4, 0.7, Inf), 
                     labels = c('Unmethylated (<0.4)', 'Intermediate (0.4-0.7)', 
                                'Methylated (>0.7)'))) %>%
  
  # count
  group_by(filt, array, Trimester, Celltype, range) %>%
  summarize(n = n())  %>%
  
  # proportion
  mutate(prop = n/sum(n),
         prop_label = percent(prop),
         
         # create x min and x max
         xmin = 
           case_when(
             grepl('<0.4', range) ~ min(densities_pmds$x)-0.0001,
             grepl('0.4-0.7', range) ~ 0.4,
             grepl('>0.7', range) ~ 0.7),
         xmax = 
           case_when(
             grepl('<0.4', range) ~ 0.4,
             grepl('0.4-0.7', range) ~ 0.7,
             grepl('>0.7', range) ~ max(densities_pmds$x)) + 0.001)

# Repeat but with more resolution:
counts_pmds_high_res <- betas_pmds %>%
  unnest(betas) %>%
  mutate(range = cut(betas, 
                     c(-Inf, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, Inf))) %>%
  
  # count
  group_by(filt, array, Trimester, Celltype, range) %>%
  summarize(n = n())  %>%
  
  # proportion
  mutate(prop = n/sum(n),
         prop_label = percent(prop)) %>%
  ungroup()

# join to densities_pmds
densities_pmds %>%
  inner_join(counts_pmds, by = c('filt', 'array', 'Trimester', 'Celltype')) %>%
  filter(x >= xmin, x <= xmax)
```

## Plot

density plots

```{r, fig.width=5.5, fig.height=7.5}
densities_pmds

densities_pmds_plots <- 
  
  # join to densities_pmds
  densities_pmds %>%
  inner_join(counts_pmds, by = c('filt', 'array', 'Trimester', 'Celltype')) %>%
  filter(x >= xmin, x <= xmax) %>%
  
  nest(density = c(Trimester:xmax)) %>%
  mutate(plots = 
           pmap(list(filt, array, density),
                ~ ggplot(data = filter(..3,  Trimester != 'Second')) +
                  
                  # density line
                    geom_line(size = 2, aes(x = x, y = y, color = Celltype)) +
                    
                  # shaded area
                    geom_area(aes(x = x, y = y, fill = Celltype, alpha = range), color = NA) +
                    
                  # percentage / AUC
                    geom_text(data = filter(..3,  Trimester != 'Second') %>%
                                 group_by(Trimester, Celltype, range) %>%
                                 summarize(x = mean(x),
                                           prop_label = prop_label[1]),
                               aes(x = x, y = -0.2, label = prop_label),
                              size = 3) +
                    theme_bw() +
                    scale_x_continuous(breaks = c(0, 0.4, 0.7, 1)) +
                    scale_y_continuous(limits = c(-0.3, NA)) +
                    facet_grid(Celltype~Trimester) +
                    scale_color_manual(values =
                                         color_code_tissue[unique(densities_pmds$Celltype)],
                                       guide = "none") +
                    scale_fill_manual(values =
                                         color_code_tissue[unique(densities_pmds$Celltype)]) +
                    scale_alpha_manual(values = c(0.1, 0.5, 0.1), guide = "none") +
                    labs(x = '% methylation', y = 'density', color = '',  fill = '',
                       title = glue('array: {..2}, cpgs: {..1}'))))

densities_pmds_plots$plots %>% print()
```

Let's look at a few specific PMDs

subset to those with a lot of cpg coverage

```{r}
# manually explore regions to plot
anno %>%
  
  filter(cpg %in% rownames(betas_filt)) %>%
  
  # count number of cpgs per pmd, parse the pmd_id into position
  group_by(pmd_id) %>%
  summarize(n = n(), pmd_width = paste(unique(pmd_width), collapse = ', ')) %>%
  filter(!is.na(pmd_id)) %>%
  separate(pmd_id, into = c('chr', 'start', 'end')) %>%
  mutate(chr = factor(chr, levels = paste0('chr', c(1:22, 'X'))),
         start = as.numeric(start),
         end = as.numeric(end)) %>%
  arrange(chr, start) %>%
  
  # filter out pmds with low coverage
  filter(n > 20) %>%
  datatable()

# define plotting function
plot_region <- function(c, s, e, span = 0.1, cpg_size = 500,
                        filtered = TRUE, array = 'EPIC') {
  
  if (filtered) {
    region1 <- anno %>%
      filter(chr == c, between(start, s, e),
             cpg %in% rownames(betas_filt),
             !cpg_id %in% c('island', 'shore'),
             !grepl('promoter', genes_id)) %>%
      pull(cpg)
  } else {
    region1 <- anno %>%
      filter(chr == c, between(start, s, e),
             cpg %in% rownames(betas_filt)) %>%
      pull(cpg)
  }
  
  if (array == '450K') {
    region1 <- intersect(anno_450k$Name, region1)
  } 
  
  # subset to cpgs in region
  region1_data <- betas_filt[region1, ] %>%
    
    t() %>%
    as.data.frame() %>%
    bind_cols(Sample_Name = rownames(.), .) %>%
    
    # melt
    gather(cpg, beta, -Sample_Name) %>%
    
    # add trimester and sex
    inner_join(pDat_filt %>% select(Sample_Name, Trimester, Tissue)) %>%
    
    # add cpg coordinates
    inner_join(anno %>% select(cpg, chr, pos = start)) %>%
    
    # set a grouping variable to control alpha for placenta/troph
    mutate(alpha_group = if_else(Tissue %in% c('Trophoblasts cs', 'Villi'),
                                 1, 0.6)) %>%
    
    # remove second trimester data
    filter(Trimester != 'Second')
  
  # cpg density for measured cpgs
  cpg_dens <- anno %>%
    filter(cpg %in% region1) %>%
    select(chr, start, end) %>%
    mutate(dens = approxfun(density(start))(start))
  
  # cpg density for all cpgs
  cpg_dens_all <- hg19_cpgs %>%
    filter(chr == c, start > s, end < e) %>%
    mutate(dens = approxfun(density(start))(start))
  
  # get pmd data
  pmd_data <- anno %>%
    filter(chr == c, between(start, s, e), 
           !is.na(pmd_id)) %>%
    select(pmd_id) %>%
    separate(pmd_id, into = c('chr', 'start', 'end'), remove = FALSE) %>%
    mutate(chr = factor(chr, levels = paste0('chr', c(1:22, 'X'))),
           start = as.numeric(start),
           end = as.numeric(end)) %>%
    arrange(chr, start) %>%
    distinct()
  
  ggplot(region1_data) +
    geom_line(stat = 'smooth', span = span, method = 'loess', se = FALSE,
              aes(x = pos, y = beta, color = Tissue, alpha = alpha_group), size = 1.25) +
    geom_rect(data = pmd_data, 
              aes(xmin = start, xmax = end), 
              ymin = 0.025, ymax = 0.05, fill = 'black') +
    geom_rect(data = cpg_dens, aes(xmin = start - cpg_size/2, xmax = start + cpg_size/2,
                                   fill = dens),
              ymin = 0, ymax = 0.025) +
    geom_rect(data = cpg_dens_all, aes(xmin = start - cpg_size/4, xmax = start + cpg_size/4,
                                   fill = dens),
              ymin = -0.025, ymax = 0) +
    facet_wrap(~Trimester, ncol = 1) +
    scale_color_manual(values= color_code_tissue[unique(region1_data$Tissue)]) +
    scale_fill_viridis_c() +
    scale_y_continuous(limits = c(0, 1.1), breaks = c(0, 0.25, 0.5, 0.75, 1)) +
    scale_x_continuous(limits = c(s, e)) +
    scale_alpha_identity() +
    labs(x = paste0(c, ':', s, '-', e), fill = 'density')
}
```

plot regions

```{r warning = FALSE, message = FALSE}
# filtered
plot_region(c = 'chr1', s = 1000000, e = 1130000, span = 0.25,filtered = T, cpg_size = 250) 

# unfiltered
plot_region(c = 'chr1', s = 1000000, e = 1130000, span = 0.25, filtered = F) 

# filtered
plot_region(c = 'chr4', s = 6400000, e = 6600000, span = 0.15, filtered = T)

# unfiltered
plot_region(c = 'chr4', s = 6400000, e = 6600000, span = 0.15, filtered = F) 


# filtered
plot_region(c = 'chr7', s = 850000, e = 1025000, span = 0.15, filtered = T)
# unfiltered
plot_region(c = 'chr7', s = 850000, e = 1025000, span = 0.15, filtered = F) 

# Figure 1d
# filtered, 450k
plot_region(c = 'chr21', s = 28000000, e = 47000000, span = 0.05, filtered = T, cpg_size = 5000,
            array = '450k')  

# filtered, EPIC
plot_region(c = 'chr21', s = 28000000, e = 47000000, span = 0.05, filtered = T, cpg_size = 5000,
            array = 'EPIC')  

# unfiltered, 450k
plot_region(c = 'chr21', s = 28000000, e = 47000000, span = 0.05, filtered = F, cpg_size = 5000,
            array = '450k')  

# unfiltered, EPIC
plot_region(c = 'chr21', s = 28000000, e = 47000000, span = 0.05, filtered = F, cpg_size = 5000,
            array = 'EPIC')  

# Figure S1c, filtered, 450k
plot_region(c = 'chr21', s = 13000000, e = 48000000, span = 0.01, filtered = T, cpg_size = 5000,
            array = '450k') +
  scale_x_continuous(limits = c(13000000, 48000000), breaks = seq(15000000, 45000000, 
                                                                  length.out = 4),
                     labels = seq(15000000, 45000000, length.out = 4) / 1000000) 

# Figure S1c, unfiltered, 450k
plot_region(c = 'chr21', s = 13000000, e = 48000000, span = 0.01, filtered = F, cpg_size = 5000,
            array = '450k') +
  scale_x_continuous(limits = c(13000000, 48000000), breaks = seq(15000000, 45000000, 
                                                                  length.out = 4),
                     labels = seq(15000000, 45000000, length.out = 4) / 1000000) 
# Figure S1c, filtered, EPIC
plot_region(c = 'chr21', s = 13000000, e = 48000000, span = 0.01, filtered = T, cpg_size = 5000,
            array = 'EPIC') +
  scale_x_continuous(limits = c(13000000, 48000000), breaks = seq(15000000, 45000000, 
                                                                  length.out = 4),
                     labels = seq(15000000, 45000000, length.out = 4) / 1000000) 
# Figure S1c, unfiltered, EPIC
plot_region(c = 'chr21', s = 13000000, e = 48000000, span = 0.01, filtered = T, cpg_size = 5000,
            array = 'EPIC') +
  scale_x_continuous(limits = c(13000000, 48000000), breaks = seq(15000000, 45000000, 
                                                                  length.out = 4),
                     labels = seq(15000000, 45000000, length.out = 4) / 1000000) 
```

# Coverage 450k/850k

Here I examine the coverage over PMDs with 450k/850k data

## Convert hg18 to hg19

```{r}
# make an identifier for original pmds
pmds <- pmds %>% 
  mutate(width = end-start,
         ID = paste0('pmd_', 1:n()))

# load chain file
chain_hg18_hg19 <- import.chain("Z:/Victor/Repositories/EPIC_annotation/hg18ToHg19.over.chain")
chain_hg18_hg19


# coerce to granges object
pmds_hg18 <- pmds %>%
  makeGRangesFromDataFrame(keep.extra.columns = T)

# hg18 to hg19 coordinates
pmds_hg19 <- liftOver(pmds_hg18, chain_hg18_hg19) %>% unlist

length(pmds_hg18); length(pmds_hg19)
```

It looks like many pmds are being broken after mapping to hg19.

```{r}
pmds_hg18 %>% 
  as_tibble() %>%
  mutate(build = 'hg18') %>%
  bind_rows(., pmds_hg19 %>% as_tibble() %>% mutate(build = 'hg19')) %>%
  ggplot(aes(x = width, fill = build)) +
  geom_histogram() +
  scale_y_log10(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0), limits = c(0, 4000000)) +
  labs(x = 'Size of PMD', fill = 'Genome build')
```

What should I do about this?

One might want to expand the remapped pmds so that they are no longer broken (reuslting in the same number of pmds before remapping). However, if the gaps between broken pieces is large, this might not make sense (because cpgs in those regions may be hypermethylated).

One could do the above merging, only when broken up pieces have small gaps (e.g. less than 500 bp). However, this would take some data manipulation.

For my question (what is the coverage of cpgs over known pmds), I think I can leave as is. CpGs in the broken gaps were not previously assessed, and therefore don't need to be included. 

Below, I
- count number of total cpgs overalapping each pmd region
- count number of EPIC cpgs overlapping each pmd region
- count number of 450k cpgs overlapping each pmd region

```{r}
# clean up
pmds_hg19 <- pmds_hg19 %>% 
  as_tibble() %>%
  select(-strand) %>%
  dplyr::rename(chr = seqnames) %>%
  mutate(chr = as.character(chr)) 

# all cpgs to pmds
pmds_hg19_cpg <- pmds_hg19 %>% 
  
  #join
  genome_left_join(hg19_cpgs %>% 
                     mutate(cpg = paste0('cpg', 1:n())),
                   
                   by = c('chr', 'start', 'end')) %>%
  select(-contains('.y')) %>%
  dplyr::rename(chr = chr.x, start = start.x, end = end.x)

# count number of cpgs per pmd
total <- pmds_hg19_cpg %>%
  group_by(ID) %>%
  summarize(Total = n(),
            width = dplyr::first(width))

# number of epic cpgs
epic <- pmds_hg19 %>% 
  genome_left_join(anno %>% 
                     filter(!cpg_id %in% c('island', 'shore'),
                            !grepl('promoter', genes_id)) %>%
                     select(chr, start, end, cpg_epic = cpg)) %>%
  select(-contains('.y')) %>%
  dplyr::rename(chr = chr.x, start = start.x, end = end.x) %>%
  dplyr::count(ID, name = 'EPIC')

k450 <- pmds_hg19 %>% 
  genome_left_join(anno %>% 
                     filter(cpg %in% anno_450k$Name,
                            !cpg_id %in% c('island', 'shore'),
                            !grepl('promoter', genes_id)) %>%
                     select(chr, start, end, cpg_epic = cpg)) %>%
  select(-contains('.y')) %>%
  dplyr::rename(chr = chr.x, start = start.x, end = end.x) %>%
  dplyr::count(ID, name = '450k')

total <- total %>% 
  
  # join data
  left_join(epic) %>%
  left_join(k450) %>%
  
  # tidy
  pivot_longer(cols = c(Total, EPIC, `450k`),
               names_to = 'count_type',
               values_to = 'n') %>%
  
  # get proportion
  group_by(ID) %>%
  mutate(prop = n/max(n)) %>%
  
  filter(count_type != 'Total', prop < 0.1)
  
# plot
total %>%
  ggplot(aes(y = prop, x = count_type)) +
  geom_boxplot() +
  scale_y_continuous(labels = percent) +
  labs(y = 'Coverage over PMDs', x = '')

total %>% 
  group_by(count_type) %>%
  summarize(median = median(prop), IQR = IQR(prop),
            mean = mean(prop), sd = sd(prop)) %>%
  mutate_if(is.numeric, percent) %>%
  kable() %>%
  kable_styling()
```

# Save

```{r, eval = FALSE}
saveRDS(densities_pmds_plots, '../../data/main/interim/2_7_pmd_density_plots.rds')
saveRDS(total, '../../data/main/interim/2_7_pmd_coverage.rds')
saveRDS(counts_pmds_high_res, '../../data/main/interim/2_7_pmd_high_res_density.rds')
```