---
title: "2_7_PMDs"
author: "Victor Yuan"
date: "23/09/2019"
output:
  html_document:
    df_print: paged
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

# Setup

## Libraries

```{r, message = FALSE, warning = FALSE}
# libraries and data
library(tidyverse); theme_set(theme_bw())
library(kableExtra)
library(glue)
```

## Data

```{r, message = FALSE, warning = FALSE}
pDat <- readRDS('../../data/main/interim/2_3_pDat_contam.rds')
pDat <- pDat %>%
  mutate(Tissue = case_when(
    !(Tissue %in% c('Villi', 'Villi maternal', 'Syncytiotrophoblast')) ~ paste(Tissue, 'cs'),
    Tissue == 'Syncytiotrophoblast' ~ 'Trophoblasts enz',
    TRUE ~ Tissue
  )) 

# raw methylation data
betas <- readRDS('../../data/main/interim/1_4_betas_noob_filt.rds')

mset_noob <- readRDS('../../data/main/interim/1_4_mset_noob.rds') # for mvals
colnames(mset_noob) <- colnames(betas) <- pDat$Sample_Name
mvals <- getM(mset_noob)

probe_anno <- readRDS('../../data/main/interim/1_1_probe_anno.rds')

# color key
pheatmap_color_code <- readRDS('../../data/main/interim/1_1_color_code.rds')

color_code <- readRDS('../../data/main/interim/2_3_color_code.rds')
color_code_tissue <- setNames(color_code$Colors_Tissue, color_code$label)

# DMCs
dmcs <- readRDS('../../data/main/interim/2_4_dmcs.rds')

# annotation
anno <- readRDS('Z:/Victor/Repositories/EPIC_annotation/hg19_epic_annotation.rds')

#color code
color_code <- readRDS('../../data/main/interim/2_3_color_code.rds')
color_code_tissue <- setNames(color_code$Colors_Tissue, color_code$label)

# 450k annotation
anno_450k <- read_csv('Z:/Victor/Data/DNAm annotations/HumanMethylation450_15017482_v1-2.csv',
                      skip = 7)

# all cpgs in the genome
hg19_cpgs <- readRDS('../../data/main/processed/0_2_all_cpgs_hg19.rds')
```

## Remove samples

remove contamined and non-interesting samples

```{r}
pDat_filt <- pDat %>% 
  filter(maternal_contamination_norm_flip < 0.35,
         !Sample_Name %in% c('PM364_hofb_cs', 'PL293_v_R2', 'PM366_vc_R2', 'P131_hofb_cs'),
         !Tissue %in% c('Villi maternal', 'Trophoblasts enz', 'Mixture cs', 'Dead Cells and Lymphocytes cs'))

# filter to first trimester
mvals_filt <- mvals[rownames(betas),pDat_filt$Sample_Name]
betas_filt <- betas[,pDat_filt$Sample_Name]
```

# PMD analysis

## Filter cpgs

D.I schroeder state that they remove probes in  the following regions:

* promoters
* cpg islands
* cpg island shores

So here I compare this filter to an unfiltered set of cpgs, and also 450k / EPIC

A table of number of cpgs per filtered/ unfiltered, and 450k/epic cpgs:

```{r}
# different sets of cpgs
cpg_sets <- expand_grid(filt = c('unfiltered', 'filtered'), 
                        array = c('450k', 'EPIC'))

# all cpgs in pmds (unfiltered, filtered)
cpg_pmd_unfilt <- anno %>%
  filter(!is.na(pmd_id),
         cpg %in% rownames(betas_filt)) 
cpg_pmd_filt <- cpg_pmd_unfilt %>%
  filter(!is.na(pmd_id),
         !cpg_id %in% c('island', 'shore'),
         !grepl('promoter', genes_id))

# get cpgs per set
cpg_sets <- cpg_sets %>%
  mutate(cpg = case_when(
    filt == 'unfiltered' & array == 'EPIC' ~ list(cpg_pmd_unfilt$cpg),
    filt == 'unfiltered' & array == '450k' ~ list(intersect(cpg_pmd_unfilt$cpg, anno_450k$IlmnID)),
    
    filt == 'filtered' & array == 'EPIC' ~ list(cpg_pmd_filt$cpg),
    filt == 'filtered' & array == '450k' ~ list(intersect(cpg_pmd_filt$cpg, anno_450k$IlmnID))
  ))

#display number of cpgs
cpg_sets %>%
  mutate(n = map_int(cpg, length)) %>%
  select(-cpg) %>%
  pivot_wider(
    id_cols = filt,
    names_from = array,
    values_from = n
  )
```

## Calculate densities

Now I can calculate the density per celltype / trimeseter. I follow the same code from my enrichment
analysis.

I create a new column in `cpg_sets` that is a nested data frame that corresponds to the density over the betas for each celltype/trimester across cpgs in each cpg set.

```{r}
# get betas in pmds, by cell type and trimester,
betas_pmds <- cpg_sets %>%
  
  # add a data-frame column that is the pDat with a nested sample names column
  mutate(pdat = list(
           pDat_filt %>%
  
           # get sample names for each group
           select(Trimester, Tissue, Sample_Name) %>%
           dplyr::rename(Celltype = Tissue) %>%
           group_by(Trimester, Celltype) %>%
           summarize(Samples =list(Sample_Name))
           )) %>%
  
  # get betas for each group
  unnest(pdat) %>%
  mutate(betas = 
           map2(cpg, Samples,
                ~ as.vector(betas_filt[.x, .y]))) %>%
  select(-cpg, -Samples)

 # calculate density
densities_pmds <- betas_pmds %>% 
  mutate(density = map(betas, density),
         x = map(density, 'x'),
         y = map(density, 'y')) %>%
  
  # clean up
  select(-betas, -density) %>%
  unnest(c(x, y))


# calculate percent methylation over regions
counts_pmds <- betas_pmds %>%
  unnest(betas) %>%
  mutate(range = cut(betas, 
                     c(-Inf, 0.4, 0.7, Inf), 
                     labels = c('Unmethylated (<0.4)', 'Intermediate (0.4-0.7)', 
                                'Methylated (>0.7)'))) %>%
  group_by(filt, array, Trimester, Celltype, range) %>%
  summarize(n = n(),
            )

counts_pmds %>%
  dplyr::count(range)
```

## Plot

density plots

```{r}
densities_pmds_plots <- densities_pmds %>%
  filter(Trimester != 'Second') %>%
  nest(c(Trimester:y)) %>%
  mutate(plots = 
           pmap(list(filt, array, data),
                ~ ggplot(data = ..3) +
                  geom_line(size = 2, aes(x = x, y = y, color = Celltype)) +
                  theme_bw() +
                  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1)) +
                  facet_grid(Trimester~Celltype) +
                  scale_color_manual(values =
                                       color_code_tissue[unique(densities_pmds$Celltype)]) +
                  labs(x = '% methylation', y = 'density', color = '', 
                       title = glue('array: {..2}, cpgs: {..1}'))))

densities_pmds_plots$plots %>% print()
```

Let's look at a few specific PMDs

subset to those with a lot of cpg coverage

```{r}
# manually explore regions to plot
anno %>%
  
  filter(cpg %in% rownames(betas_filt)) %>%
  
  # count number of cpgs per pmd, parse the pmd_id into position
  group_by(pmd_id) %>%
  summarize(n = n(), pmd_width = paste(unique(pmd_width), collapse = ', ')) %>%
  filter(!is.na(pmd_id)) %>%
  separate(pmd_id, into = c('chr', 'start', 'end')) %>%
  mutate(chr = factor(chr, levels = paste0('chr', c(1:22, 'X'))),
         start = as.numeric(start),
         end = as.numeric(end)) %>%
  arrange(chr, start) %>%
  
  # filter out pmds with low coverage
  filter(n > 20) %>%
  datatable()

# define plotting function
plot_region <- function(c, s, e, span = 0.1, cpg_size = 500,
                        filtered = TRUE, array = 'EPIC') {
  
  if (filtered) {
    region1 <- anno %>%
      filter(chr == c, between(start, s, e),
             cpg %in% rownames(betas_filt),
             !cpg_id %in% c('island', 'shore'),
             !grepl('promoter', genes_id)) %>%
      pull(cpg)
  } else {
    region1 <- anno %>%
      filter(chr == c, between(start, s, e),
             cpg %in% rownames(betas_filt)) %>%
      pull(cpg)
  }
  
  if (array == '450K') {
    region1 <- intersect(anno_450k$Name, region1)
  } 
  
  # subset to cpgs in region
  region1_data <- betas_filt[region1, ] %>%
    
    t() %>%
    as.data.frame() %>%
    bind_cols(Sample_Name = rownames(.), .) %>%
    
    # melt
    gather(cpg, beta, -Sample_Name) %>%
    
    # add trimester and sex
    inner_join(pDat_filt %>% select(Sample_Name, Trimester, Tissue)) %>%
    
    # add cpg coordinates
    inner_join(anno %>% select(cpg, chr, pos = start)) %>%
    
    # set a grouping variable to control alpha for placenta/troph
    mutate(alpha_group = if_else(Tissue %in% c('Trophoblasts cs', 'Villi'),
                                 1, 0.6)) %>%
    
    # remove second trimester data
    filter(Trimester != 'Second')
  
  # cpg density for measured cpgs
  cpg_dens <- anno %>%
    filter(cpg %in% region1) %>%
    select(chr, start, end) %>%
    mutate(dens = approxfun(density(start))(start))
  
  # cpg density for all cpgs
  cpg_dens_all <- hg19_cpgs %>%
    filter(chr == c, start > s, end < e) %>%
    mutate(dens = approxfun(density(start))(start))
  
  # get pmd data
  pmd_data <- anno %>%
    filter(chr == c, between(start, s, e), 
           !is.na(pmd_id)) %>%
    select(pmd_id) %>%
    separate(pmd_id, into = c('chr', 'start', 'end'), remove = FALSE) %>%
    mutate(chr = factor(chr, levels = paste0('chr', c(1:22, 'X'))),
           start = as.numeric(start),
           end = as.numeric(end)) %>%
    arrange(chr, start) %>%
    distinct()
  
  ggplot(region1_data) +
    geom_line(stat = 'smooth', span = span, method = 'loess', se = FALSE,
              aes(x = pos, y = beta, color = Tissue, alpha = alpha_group), size = 1.25) +
    geom_rect(data = pmd_data, 
              aes(xmin = start, xmax = end), 
              ymin = 0.025, ymax = 0.05, fill = 'black') +
    geom_rect(data = cpg_dens, aes(xmin = start - cpg_size/2, xmax = start + cpg_size/2,
                                   fill = dens),
              ymin = 0, ymax = 0.025) +
    geom_rect(data = cpg_dens_all, aes(xmin = start - cpg_size/4, xmax = start + cpg_size/4,
                                   fill = dens),
              ymin = -0.025, ymax = 0) +
    facet_wrap(~Trimester, ncol = 1) +
    scale_color_manual(values= color_code_tissue[unique(region1_data$Tissue)]) +
    scale_fill_viridis_c() +
    scale_y_continuous(limits = c(0, 1.1), breaks = c(0, 0.25, 0.5, 0.75, 1)) +
    scale_x_continuous(limits = c(s, e)) +
    scale_alpha_identity() +
    labs(x = paste0(c, ':', s, '-', e), fill = 'density')
}
```

plot regions

```{r warning = FALSE, message = FALSE}
plot_region(c = 'chr1', s = 1000000, e = 1130000, span = 0.25,filtered = T, cpg_size = 250) 
plot_region(c = 'chr1', s = 1000000, e = 1130000, span = 0.25, filtered = F) 

plot_region(c = 'chr4', s = 6400000, e = 6600000, span = 0.15, filtered = T)
plot_region(c = 'chr4', s = 6400000, e = 6600000, span = 0.15, filtered = F) 

plot_region(c = 'chr7', s = 850000, e = 1025000, span = 0.15, filtered = T)
plot_region(c = 'chr7', s = 850000, e = 1025000, span = 0.15, filtered = F) 

# Figure 1d
plot_region(c = 'chr21', s = 28000000, e = 47000000, span = 0.05, filtered = T, cpg_size = 5000,
            array = '450k')  
plot_region(c = 'chr21', s = 28000000, e = 47000000, span = 0.05, filtered = T, cpg_size = 5000,
            array = 'EPIC')  

# Figure S1c
plot_region(c = 'chr21', s = 13000000, e = 48000000, span = 0.01, filtered = T, cpg_size = 5000,
            array = '450k') +
  scale_x_continuous(limits = c(13000000, 48000000), breaks = seq(15000000, 45000000, 
                                                                  length.out = 4),
                     labels = seq(15000000, 45000000, length.out = 4) / 1000000) 
```
