---
title: "01_Load_IDATs"
author: "Victor Yuan"
date: "April 27, 2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

Load week 1&2 samples (n=192, 24 chips)

# 1.0 Libraries

```{r}
library(tidyr)
library(plyr)
library(dplyr)
library(minfi)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(IlluminaHumanMethylationEPICmanifest)
library(readxl)
library(ewastools)
```

# 2.0 RGSET

```{r}
#sample phenodata 
ss <- read_xlsx('../../data/main/raw/NIH EPIC Batch 2 sample sheet.xlsx', skip = 7)
ss

# create sentrix column which matches column names of rgset
ss <- ss %>% 
  mutate(Sentrix = paste0(Sentrix_ID, '_', Sentrix_Position)) %>%
  dplyr::rename(Chip_number = `Chip Number`,
                Scratches = `Scratches on beadchip`,
                DNA_conc_BSC_adjusted = `BSC sample adjustedconcentration`,
                DNA_conc_before_load = `DNA concentration after speed vac and dilution (ng/ul)`,
                DNA_loaded = `Amount of DNA loaded (ng)`) %>%
  arrange(Sentrix)
```

```{r}
# load idats
rgset <- read.metharray.exp('Z:/ROBLAB6 Infinium450k John/EPIC Raw data/NIH EPIC Batch 2/', 
                            recursive = T, verbose = T, extended = T)
rgset

sum(ss$Sentrix %in% colnames(rgset)) #192

#rearrange argset based on ss
rgset <- rgset[,ss$Sentrix]

# add pData to rgset
ss1 <- ss %>% as.data.frame()
rownames(ss1) <- ss1$Sentrix
pData(rgset) <- DataFrame(ss1)
pData(rgset)
```

```{r, eval = F}
saveRDS(rgset, '../../data/main/interim/01_rgset_raw.rds')
```

# 3.0 ewastools

Here I construct a ewastools -loaded object to use their functions

```{r}
# construct filepaths
Base <- 'Z:/ROBLAB6 Infinium450k John/EPIC Raw data/NIH EPIC Batch 2/'
week1 <- 'IDATs - Week 1/'
week2 <- 'IDATs - Week 2/'
  
ss <- ss %>% 
  mutate(filenames = ifelse(Week == 1, 
                            paste0(Base, week1, Sentrix_ID, '/', Sentrix),
                            paste0(Base, week2, Sentrix_ID, '/', Sentrix)))

et_meth <- read_idats(ss$filenames) # `quiet=TRUE` supresses the progress bar, 5.5 Gb
```

Note that et_meth is a list containing the following elements:

manifest
M - methylated matrix
U - unmethylated matrix
N - number of beads for M
V - number of beads for U
ctrlG - intensities for control probes in green
ctrlR - intensities for ctrl probes in red
meta - metadata for samples


Do not ever try to print the object to screen. (way to large).

saveRDS(et_meth, '../../data/main/interim/01_ewastools_loaded_idats.rds')

# Session Info

```{r}
sessionInfo()
```