---
title: "0_2_download_cpgs"
author: "Victor Yuan"
date: "24/09/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

This script is for downloading all cpgs of the genome. The purpose is so I can see how many cpgs 
overlap with EPIC for pmd regions.

# Libraries

```{r, message = F, warning = F}
library(BSgenome.Hsapiens.UCSC.hg19)
library(Biostrings)
library(tidyverse)
library(fuzzyjoin)
```

# Find all cpgs in the genome

```{r}
genome <- BSgenome.Hsapiens.UCSC.hg19

# target chromosomes
seqnames(genome)[1:24] 

# example counting cgs in chr 1
countPattern(pattern = DNAString('cg'), genome$chr1)

# target pattern
pat <- DNAStringSet(list(cpg = DNAString('cg')))

# find pattern across each chromosome
results <- tibble(chr = seqnames(genome)[1:24]) %>%
  
  mutate(matches = map(chr, ~ extractAllMatches(genome[[.x]],
                                                matchPDict(pat, genome[[.x]]))),
         
         # pull out start and end positions
         start = map(matches, ~start(.x)),
         end = map(matches, ~end(.x)))


results <- results %>%
  select(-matches) %>%
  unnest()

results
```

28 million cpgs, that's what we expect

Let's verify that the EPIC cpgs are represented in our results

```{r}
EPIC <- readRDS('Z:/Victor/Repositories/EPIC_annotation/hg19_epic_annotation.rds')
results %>%
  genome_inner_join(EPIC %>% 
                      select(cpg, chr, start, end),
                    c('chr', 'start', 'end'))
nrow(EPIC)

EPIC %>% 
  select(cpg, chr, start, end) %>%
  genome_anti_join(results, c('chr', 'start', 'end')) %>%
  tail

```

Great! I found all cpgs in the genome, and they match up with the EPIC cpgs. The EPIC annotation also has non-cpg sites, which are obviously not included in the results. So that all makes sense.


# Save results

```{r, eval = F}
saveRDS(results, '../../data/main/processed/0_2_all_cpgs_hg19.rds')
```
