---
title: "5_1_dbGap_prepare_files"
author: "Victor Yuan"
date: "09/04/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

here I prepare files for dbgap submission

# libraries

```{r}
library(here)
library(tidyverse)
```

# phenotype data

```{r}
base_path <- file.path('data', 'main', 'interim')

# pData
pDat <- readRDS(here(base_path, '2_3_pDat_contam.rds'))

pDat <- pDat %>%
  mutate(Tissue = case_when(
    !(Tissue %in% c('Villi', 'Villi maternal', 'Syncytiotrophoblast', 'Mixture')) ~ paste(Tissue, 'cs'),
    Tissue == 'Syncytiotrophoblast' ~ 'Trophoblasts enz',
    TRUE ~ Tissue
  )) 

# filter out samples
pDat_filt <- pDat %>% 
  filter(maternal_contamination_norm_flip < 0.35,
         !Sample_Name %in% c('PM364_hofb_cs', 'PL293_v_R2', 'PM366_vc_R2', 'P131_hofb_cs', 
                             'PM324_V4', 'PM324_V1', 'PM139_vc', 'PM77_vc'),
         !Tissue %in% c('Villi maternal', 
                        'Dead Cells and Lymphocytes cs',
                        'Mixture'),
         Trimester != 'Second')

pDat_filt %>% count(Case_ID) %>% as.data.frame
pDat_filt %>% count(Tissue) %>% as.data.frame

# filter out columns
pDat_filt %>%
  select(Sample_Name, Case_ID, 
         Sex, GA, Trimester, Tissue, 
         Sentrix_ID, Sentrix_Position, Chip_number, Well) %>%
  rename(Celltype = Tissue, Gestational_age = GA)
```

# Create deidentified IDs

```{r}
#subject IDs
sub_id <-pDat_filt %>%
  group_by(Case_ID) %>%
  summarize %>%
  mutate(SUBJECT_ID = as.character(1001:(1000+n())))

pDat_filt <- pDat_filt %>%
  arrange(Case_ID) %>%
  left_join(sub_id) %>%
  mutate(SAMPLE_ID = paste0('S', 1:n()))

pDat_filt %>%
  select(Case_ID, Sample_Name, SUBJECT_ID, SAMPLE_ID)  %>%
  write_delim(here('dbGap submission', '0_roblab_to_dbGap_sample_id_mapping.txt'), delim = '\t')
```



# 2a_SubjectConsent_DS.txt and 2b_SubjectConsent_DD.xlsx

**Subject consent DS** must contain columns:

- SUBJECT_ID, unique deidentified identifier for each individual	
- CONSENT


```{r}
pDat_filt %>% 
  group_by(SUBJECT_ID) %>% 
  summarize(CONSENT = 1) %>% 
  write_delim(here('dbGap submission', '2a_SubjectConsent_DS.txt'), delim = '\t')
```

# 3a_SSM_DS

subject sample mapping 

- SUBJECT_ID
- SAMPLE_ID

```{r}
pDat_filt %>% 
  select(SUBJECT_ID, SAMPLE_ID) %>%
  arrange(SUBJECT_ID, SAMPLE_ID) %>%
  write_delim(here('dbGap submission', '3a_SSM_DS.txt'), delim = '\t')
```

#5a_SubjectPhenotypes_DS.txt

- SUBJECT_ID
- phenotype variables

## add redcap ga info

```{r}
rc <- read_csv(here('data', 'main', 'raw',
                    'RobLabPMCaseDatabase-Victorcellspmsamples_DATA_2020-03-18_1502.csv'))
rc_filt <- rc %>% 
  filter(study_id %in% pDat_filt$Case_ID)

rc_filt %>% filter(is.na(del_ga))

rc_filt %>%
  filter(!is.na(del_ga),
         !study_id %in% c('PM77','PM139')) %>%
  summarize(ga_mean = mean(del_ga),
            ga_min = min(del_ga),
            ga_max = max(del_ga),
            ga_sd = sd(del_ga))
```

```{r}
ethnicity <- rc_filt %>%
  select(study_id, contains('eth')) %>%
  pivot_longer(cols = -c(study_id,  eth_mother_type, eth_father_type),
               names_to = 'eth',
               values_to = 'yes') %>%
  mutate(ethnicity = case_when(
    grepl('0$', eth) ~ 'Caucasian',
    grepl('1$', eth) ~ 'East Asian',
    grepl('2$', eth) ~ 'South Asian',
    grepl('3$', eth) ~ 'First Nation',
    grepl('4$', eth) ~ 'Middle Eastern',
    grepl('5$', eth) ~ 'Latin American',
    grepl('6$', eth) ~ 'African',
    grepl('7$', eth) ~ 'Mixed Ethnicity',
    grepl('98$', eth) ~ 'Other',
    grepl('99$', eth) ~ 'Unknown',
    
  )) %>%
  mutate(parent = if_else(grepl('mother', eth), 'eth_mother', 'eth_father')) %>%
  filter(yes == 1) %>%
  select(-eth, -yes) %>%
  pivot_wider(id_cols = c(study_id, contains('type')),
              names_from = 'parent',
              values_from = 'ethnicity',
              values_fn = list(ethnicity = function(x)paste0(x, collapse = ', '))) %>% # View here
  
  # remove extraneous info, summarize child's ethnicity
  select(-contains('type')) %>%
  mutate(ETHNICITY = ifelse(eth_mother != eth_father, 
                            paste0(eth_mother, '/', eth_father),
                            eth_mother))%>%
  
  # save
  rename(Case_ID = study_id) %>%
  select(Case_ID, ETHNICITY)


pDat_filt %>%
  select(Case_ID, SUBJECT_ID, Sex, Trimester) %>%
  rename(SEX = Sex,
         GESTATIONAL_AGE_CAT = Trimester) %>%
  mutate(GESTATIONAL_AGE_CAT = ifelse(GESTATIONAL_AGE_CAT == 'Third', 'Term', GESTATIONAL_AGE_CAT)) %>%
  unique() %>%
  arrange(SUBJECT_ID) %>% 
  left_join(ethnicity) %>%
  select(-Case_ID) %>%
  write_delim(here('dbGap submission', '5a_SubjectPhenotypes_DS.txt'), delim = '\t')

```

# 6a_SampleAttributes_DS.txt

- SAMPLE_ID
- BODY_SITE
- ANALYTE_TYPE
- HISTOLOGICAL_TYPE
- IS_TUMOUR

```{r}
pDat_filt %>%
  select(SAMPLE_ID, Tissue) %>%
  rename(HISTOLOGICAL_TYPE = Tissue) %>%
  mutate(BODY_SITE = 'Placenta',
         HISTOLOGICAL_TYPE = case_when(
           HISTOLOGICAL_TYPE == 'Villi' ~ 'Chorionic Villi',
           HISTOLOGICAL_TYPE == 'Endothelial cs' ~ 'Endothelial Cells',
           HISTOLOGICAL_TYPE == 'Stromal cs' ~ 'Stromal Cells',
           HISTOLOGICAL_TYPE == 'Trophoblasts cs' ~ 'Trophoblasts',
           HISTOLOGICAL_TYPE == 'Hofbauer cs' ~ 'Hofbauer Cells'
         ),
         ANALYTE_TYPE = 'DNA',
         IS_TUMOUR = 'no') %>%
  write_delim(here('dbGap submission', '6a_SampleAttributes_DS.txt'), delim = '\t')
```

# processed methylation data

```{r}
# processed methylation data
betas <- readRDS(here(base_path, '1_4_betas_noob_filt.rds'))
dim(betas)
betas_filt <- betas[,pDat_filt$Sentrix]
dim(betas_filt)

colnames(betas_filt) <- pDat_filt$SAMPLE_ID

write_tsv(as.data.frame(betas_filt), here('dbGap submission', 'processed_methylation_data.tsv'))
```

# raw methylation data

move idats of the 131 samples to raw data folder

## FSM

create file sample mapping 

```{r}
FSM <- pDat_filt %>%
  select(SUBJECT_ID, SAMPLE_ID, Sentrix) %>%
  mutate(FILE1 = paste0(Sentrix, '_Grn.idat'),
         FILE2 = paste0(Sentrix, '_Red.idat')) %>%
  select(-Sentrix) 

FSM %>%
  write_delim(here('dbGap submission', 'methylation data', 'raw_data_FSM.txt'), delim ='\t')
```

## copy files over

```{r}
filepaths <- 
  
  tibble(filepath = list.files(path = 'Z:/ROBLAB6 InfiniumSequenom/EPIC Raw data/NIH EPIC Batch 2',
                             recursive = TRUE, full.names = TRUE)) %>%
  filter(grepl('.idat', filepath)) %>%
  
  # get basename
  mutate(Sentrix = str_extract(filepath, '[0-9]+_[A-z0-9]{1,6}(?=_[GrnRed]{3}\\.idat)')) %>%
   
   # remove duplicates
   distinct() %>%
  
  filter(Sentrix %in% pDat_filt$Sentrix) %>%
  select(Sentrix, everything()) %>%
  
  # make destination filepaths
  mutate(destination = paste0(here('dbGap submission', 'raw data', basename(filepath))))

file.copy(filepaths$filepath, filepaths$destination, overwrite = TRUE, recursive = TRUE)
```




